---
title: "HAV Batch Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
    self_contained: true
params:
  batch_dir:    "data/Batch-1"
  batch_fa:     "data/Batch-1/Batch-1.fa"
  dataset_date: "2026-04-10"
  dataset_dir:  "data/local_dataset/2026-04-10"
  out_base:     "data/Batch-1/output"
  n_neighbors:  30
  mode:         "sanger"
  zoom_count: 5
  zoom_width: 35
  min_called_fraction: 0.5
  entropy_window: 15
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 7
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ape)
  library(ggtree)
  library(phangorn)
  library(kableExtra)
  library(scales)
  library(patchwork)
  library(jsonlite)
  library(pheatmap)
  library(grid)
})

# ── Paths ────────────────────────────────────────────────────────────────────
proj_dir     <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
                              mustWork = FALSE)
#batch_dir    <- file.path(proj_dir, params$batch_dir)
batch_dir    <- file.path(params$batch_dir)
#out_dir      <- file.path(batch_dir, "output")
out_dir      <- file.path(params$out_base)
trees_dir    <- file.path(out_dir, "trees")
#dataset_base <- file.path(proj_dir, "data", "local_dataset", params$dataset_date)
dataset_base <- file.path(params$dataset_dir)
metadata_tsv <- file.path(dataset_base, "metadata.tsv")
batch_name   <- basename(params$batch_dir)
input_fasta  <- file.path(params$batch_fa)

# ── Helper: safe read_tsv ────────────────────────────────────────────────────
safe_tsv <- function(path, ...) {
  if (!file.exists(path)) return(tibble())
  read_tsv(path, show_col_types = FALSE, ...)
}

safe_csv_sc <- function(path, ...) {
  # NextClade CSV uses semicolons
  if (!file.exists(path)) return(tibble())
  read_delim(path, delim = ";", show_col_types = FALSE, ...)
}

# Safe FASTA reader with validation (prevents ape::read.dna segfault)
safe_read_dna_fasta <- function(path) {
  # Check file existence
  if (!file.exists(path)) return(NULL)
  
  # Check file size (empty file would cause problems)
  if (file.size(path) == 0) return(NULL)
  
  # Quick validation: first line should be a header starting with '>'
  first_line <- readLines(path, n = 1)
  if (length(first_line) == 0 || !grepl("^>", first_line)) {
    warning("Invalid FASTA file (no valid header): ", path)
    return(NULL)
  }
  
  # Count sequences by counting '>' lines
  all_lines <- readLines(path)
  n_seqs <- sum(grepl("^>", all_lines))
  if (n_seqs < 1) return(NULL)
  
  # Try to read with ape::read.dna, but catch errors
  # (tryCatch won't catch segfaults, but at least we tried validation)
  tryCatch({
    ape::read.dna(path, format = "fasta", as.character = TRUE)
  }, error = function(e) {
    warning("Error reading FASTA file: ", path, " - ", e$message)
    NULL
  })
}

# Null-coalescing helper (used in parse_nearest)
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

# Normalize IDs to match FASTA headers: spaces → "_", non-ASCII stripped
norm_id <- function(x) {
  # Repair invalid byte sequences first (source TSVs are not always valid UTF-8)
  x <- iconv(x, from = "UTF-8", to = "UTF-8", sub = "")
  x <- gsub(" ", "_", x)
  iconv(x, from = "UTF-8", to = "ASCII", sub = "")
}

# ── Load data ─────────────────────────────────────────────────────────────────

# Reference metadata
meta <- safe_tsv(metadata_tsv)

#if (nrow(meta) > 0 && "date" %in% names(meta)) {
#  meta <- meta %>% mutate(date = as.character(date))
#}

if (params$dataset_date == "2026-04-10" && nrow(meta) > 0 && "id" %in% names(meta)) {
  meta <- meta %>% filter(as.character(id) != "534954")
}

# Community sequence metadata from hav-vp1-2b-lineages tree.json
# Provides clade_membership + lineage_phylo for NCBI accessions used as neighbors
.ltree_json <- file.path(proj_dir, "data", "nextclade_datasets",
                         "hav-vp1-2b-lineages", "tree.json")
community_meta <- tibble(id = character(),
                         community_clade = character(),
                         community_lineage_phylo = character())
if (file.exists(.ltree_json)) {
  .ltree <- fromJSON(.ltree_json, simplifyDataFrame = FALSE)
  .cl_leaves <- function(node) {
    if (is.null(node$children) || length(node$children) == 0) {
      na <- node$node_attrs %||% list()
      list(tibble(
        id                      = node$name %||% NA_character_,
        community_clade         = na$clade_membership$value %||% NA_character_,
        community_lineage_phylo = na$lineage_phylo$value    %||% NA_character_
      ))
    } else {
      unlist(lapply(node$children, .cl_leaves), recursive = FALSE)
    }
  }
  community_meta <- bind_rows(.cl_leaves(.ltree$tree)) %>%
    mutate(
      community_clade         = str_to_upper(community_clade),
      community_clade         = if_else(is.na(community_clade) | community_clade %in% c("", "UNASSIGNED"),
                                        NA_character_, community_clade),
      community_lineage_phylo = if_else(is.na(community_lineage_phylo) | community_lineage_phylo %in% c("", "unassigned"),
                                        NA_character_, community_lineage_phylo)
    )
  if (params$dataset_date == "2026-04-10") {
    community_meta <- community_meta %>% filter(as.character(id) != "534954")
  }
  rm(.ltree, .cl_leaves)
}
rm(.ltree_json)

# BLAST results
blast_raw <- safe_tsv(file.path(out_dir, "blast_results.tsv"))
if (params$dataset_date == "2026-04-10" && nrow(blast_raw) > 0 && "sseqid" %in% names(blast_raw)) {
  blast_raw <- blast_raw %>% filter(as.character(sseqid) != "534954")
}

# ── Amplicon coverage + primer scheme (whole-genome / ViroConstrictor only) ───
# ViroConstrictor writes one mean-coverage value per amplicon per sample to
# all_amplicon_coverage.csv. Sanger batches have no such file, so the whole
# amplicon section is skipped downstream when `has_amplicon` is FALSE.
amplicon_cov_path <- file.path(batch_dir, "viroconstrictor", "results",
                               "combined", "all_samples", "all_amplicon_coverage.csv")
has_amplicon  <- file.exists(amplicon_cov_path)
amplicon_long <- tibble()
mincov_val    <- NA_real_

if (has_amplicon) {
  .ac <- readr::read_csv(amplicon_cov_path, show_col_types = FALSE)
  # Columns: Virus, Reference_ID, amplicon_names (= sample name), then one
  # column per amplicon (e.g. HAVIB_001 … HAVIB_006).
  .amp_cols <- setdiff(names(.ac), c("Virus", "Reference_ID", "amplicon_names"))
  if (length(.amp_cols) > 0) {
    amplicon_long <- .ac %>%
      rename(sample = amplicon_names) %>%
      pivot_longer(all_of(.amp_cols), names_to = "amplicon_raw", values_to = "coverage") %>%
      mutate(
        sample   = word(as.character(sample), 1),
        amplicon = as.integer(str_extract(amplicon_raw, "[0-9]+$")),
        coverage = as.numeric(coverage)
      ) %>%
      arrange(sample, amplicon)
  }
}

# Primer scheme table (primer name ↔ amplicon number) from the batch samplesheet
primer_scheme_tbl <- tibble()
.samplesheet_path <- file.path(batch_dir, "samplesheet.tsv")
if (has_amplicon && file.exists(.samplesheet_path)) {
  .ss <- safe_tsv(.samplesheet_path)
  if ("MinCov" %in% names(.ss)) {
    mincov_val <- suppressWarnings(as.numeric(.ss$MinCov[1]))
  }
  if ("Primers" %in% names(.ss)) {
    .primer_rel   <- unique(.ss$Primers)
    .primer_rel   <- .primer_rel[!is.na(.primer_rel) & .primer_rel != "NONE"]
    .primer_paths <- file.path(proj_dir, .primer_rel)
    .primer_paths <- .primer_paths[file.exists(.primer_paths)]
    if (length(.primer_paths) > 0) {
      # Read primer records (header + sequence) preserving file order.
      .read_primer_fa <- function(p) {
        ln  <- readLines(p, warn = FALSE)
        hdr <- which(startsWith(ln, ">"))
        tibble(primer   = sub("^>", "", ln[hdr]),
               sequence = toupper(ln[pmin(hdr + 1L, length(ln))]))
      }
      .renamed_fa <- bind_rows(lapply(.primer_paths, .read_primer_fa))

      # The original (pre-ViroConstrictor) primer names live in a sibling
      # *_original.fasta with identical record order, so pair them positionally.
      .orig_paths <- sub("_renamed\\.fasta$", "_original.fasta", .primer_paths)
      .orig_paths <- .orig_paths[file.exists(.orig_paths)]
      .renamed_fa$original <- NA_character_
      if (length(.orig_paths) > 0) {
        .orig_fa <- bind_rows(lapply(.orig_paths, .read_primer_fa))
        if (nrow(.orig_fa) == nrow(.renamed_fa))
          .renamed_fa$original <- .orig_fa$primer
      }

      # Primer names look like HAVIB_1_LEFT / HAVIB_1_RIGHT — scheme, amplicon, side.
      primer_scheme_tbl <- .renamed_fa %>%
        distinct(primer, .keep_all = TRUE) %>%
        tidyr::extract(primer, c("scheme", "amplicon", "side"),
                       "^(.*)_([0-9]+)_(LEFT|RIGHT)$", remove = FALSE) %>%
        filter(!is.na(amplicon)) %>%
        mutate(amplicon = as.integer(amplicon)) %>%
        select(scheme, amplicon, side, primer, original, sequence) %>%
        pivot_wider(names_from = side,
                    values_from = c(primer, original, sequence)) %>%
        arrange(scheme, amplicon)
    }
  }
}

# ── Genotype helpers ──────────────────────────────────────────────────────────
# HAV genotypes are written in canonical roman form with an uppercase sub-letter
# (IA, IB, IIA, IIIA). Two normalisations are needed:
#   * harmonise_genotype()  — upper-case + trim any genotype label (Ib → IB).
#   * genotype_from_name()  — derive a genotype from a reference name that
#     encodes it in arabic form (e.g. "helgenom_gt_1a" → IA) for hits that carry
#     no genotype in the metadata.
.arabic_to_roman_gt <- c(`1` = "I", `2` = "II", `3` = "III",
                         `4` = "IV", `5` = "V", `6` = "VI", `7` = "VII")

harmonise_genotype <- function(x) {
  x <- str_to_upper(str_trim(as.character(x)))
  x[x %in% c("", "NA")] <- NA_character_
  x
}

genotype_from_name <- function(x) {
  m     <- str_match(str_to_lower(as.character(x)), "gt[_-]?([0-9]+)\\s*([a-z]?)")
  roman <- .arabic_to_roman_gt[m[, 2]]
  ifelse(is.na(roman), NA_character_, paste0(roman, str_to_upper(m[, 3])))
}

# NextClade — lineages dataset (clade + lineage_phylo)
nc_lineages_raw <- safe_tsv(file.path(out_dir, "lineages", "nextclade.tsv"))

# ── Check for NextClade alignment failures ──────────────────────────────────
nc_errors <- tibble()
if (nrow(nc_lineages_raw) > 0 && "errors" %in% names(nc_lineages_raw)) {
  nc_errors <- nc_lineages_raw %>%
    filter(!is.na(errors) & errors != "") %>%
    select(seqName, errors)
  if (nrow(nc_errors) > 0) {
    warning(
      "NextClade alignment failed for ", nrow(nc_errors), " sequence(s).\n",
      "This usually means sequences are too short or don't match the reference dataset.\n",
      "Try running with --min-match-length 10 or lower (default is 10)."
    )
  }
}

nc_lineages <- nc_lineages_raw
if (nrow(nc_lineages) > 0) {
  if (!"lineage_phylo" %in% names(nc_lineages)) {
    nc_lineages <- nc_lineages %>% mutate(lineage_phylo = NA_character_)
  }
  nc_lineages <- nc_lineages %>%
    select(
      seqName,
      clade,
      lineage_phylo,
      totalSubstitutions,
      coverage,
      qc_status = qc.overallStatus,
      qc_score  = qc.overallScore,
      substitutions,
      aaSubstitutions,
      privateNucMuts_total = privateNucMutations.totalPrivateSubstitutions
    ) %>%
    # Ensure coverage is numeric (NextClade may return FALSE for failed alignments)
    mutate(coverage = as.numeric(coverage))
} else {
  nc_lineages <- tibble(
    seqName = character(), clade = character(), lineage_phylo = character(),
    totalSubstitutions = integer(), coverage = double(), qc_status = character(),
    qc_score = double(), substitutions = character(), aaSubstitutions = character(),
    privateNucMuts_total = integer()
  )
}

# Harmonise the NextClade clade label to canonical uppercase (IA, IB, IIIA …).
if (nrow(nc_lineages) > 0) {
  nc_lineages <- nc_lineages %>% mutate(clade = harmonise_genotype(clade))
}

# All query names (from NC-lineages)
all_queries <- nc_lineages$seqName

# Metadata for joining reference info to BLAST hits
meta_base <- meta %>%
  transmute(
    sseqid = id,
    hit_genotype = if ("genotype" %in% names(meta)) as.character(genotype) else NA_character_,
    hit_variant = if (!is.na(meta_lineage_col)) as.character(.data[[meta_lineage_col]]) else NA_character_
  )

meta_base
} else {
  tibble(sseqid = character())
}


# Only proceed with join if blast_raw has data and the required column exists
blast_top <- if (nrow(blast_raw) > 0 && "sseqid" %in% names(blast_raw)) {
  blast_raw %>%
    left_join(meta_for_join, by = "sseqid") %>%
    # Define "nearest BLAST hit" as lowest SNP mismatch first, then highest bitscore.
    arrange(qseqid, mismatch, desc(bitscore)) %>%
    group_by(qseqid) %>%
    mutate(hit_rank = row_number()) %>%
    ungroup()
} else {
  # Return empty tibble with expected structure if no BLAST results
  blast_raw %>% mutate(hit_rank = NA_integer_)
}

blast_best <- blast_top %>%
  filter(hit_rank == 1) %>%
  select(qseqid, sseqid,
         blast_pident    = pident,
         blast_mismatch  = mismatch,
         blast_qcovs     = qcovs,
         blast_bitscore  = bitscore,
         any_of(c("hit_variant", "hit_genotype")))

# Ensure hit_variant and hit_genotype columns exist
if (!"hit_variant" %in% names(blast_best)) blast_best$hit_variant <- NA_character_
if (!"hit_genotype" %in% names(blast_best)) blast_best$hit_genotype <- NA_character_

# Harmonise and rename
blast_best <- blast_best %>%
  mutate(
    blast_variant  = hit_variant,
    blast_genotype = coalesce(as.character(harmonise_genotype(hit_genotype)),
                              as.character(genotype_from_name(sseqid)))
  ) %>%
  select(-any_of(c("hit_variant", "hit_genotype")))

# ── Master summary table ──────────────────────────────────────────────────────
# BLAST truncates FASTA headers at the first whitespace, so its qseqid is only
# the first token of the full sequence name (e.g. "barcode84" rather than
# "barcode84 HAV LC128713.1_IB mincov=20"). Join on that first token so WGS
# consensus headers (which carry spaces) still match. For single-word Sanger
# names the first token equals the full name, so this is backward-compatible.
summary_tbl <- nc_lineages %>%
  mutate(.join_key = word(seqName, 1)) %>%
  left_join(blast_best %>% 
            mutate(.join_key = word(qseqid, 1)) %>% 
            select(.join_key, qseqid, everything()),
            by = ".join_key") %>%
  select(-.join_key)

# ── Input FASTA summary (raw batch input) ───────────────────────────────────
read_fasta_simple <- function(path) {
  if (!file.exists(path)) return(tibble(header = character(), sequence = character()))

  lines <- readLines(path, warn = FALSE)
  if (length(lines) == 0) return(tibble(header = character(), sequence = character()))

  recs <- list()
  cur_h <- NULL
  cur_s <- character()

  flush_rec <- function() {
    if (is.null(cur_h)) return(NULL)
    recs[[length(recs) + 1]] <<- tibble(
      header = cur_h,
      sequence = paste(cur_s, collapse = "")
    )
    NULL
  }

  for (ln in lines) {
    ln <- trimws(ln)
    if (!nzchar(ln)) next
    if (startsWith(ln, ">")) {
      flush_rec()
      cur_h <- sub("^>", "", ln)
      cur_s <- character()
    } else {
      cur_s <- c(cur_s, ln)
    }
  }
  flush_rec()

  bind_rows(recs)
}

parse_primer_header <- function(header) {
  parts <- strsplit(header, "[[:space:]]+")[[1]]
  name <- parts[1]
  meta <- list()
  if (length(parts) > 1) {
    for (token in parts[-1]) {
      if (grepl("=", token, fixed = TRUE)) {
        kv <- strsplit(token, "=", fixed = TRUE)[[1]]
        if (length(kv) == 2) meta[[tolower(kv[1])]] <- kv[2]
      }
    }
  }
  list(name = name, meta = meta)
}

iupac_sets <- c(
  A = "A", C = "C", G = "G", T = "T", U = "T",
  R = "AG", Y = "CT", S = "GC", W = "AT", K = "GT", M = "AC",
  B = "CGT", D = "AGT", H = "ACT", V = "ACG", N = "ACGT", I = "ACGT", `5` = "ACGT"
)

primer_rc <- function(seq) {
  seq <- str_to_upper(seq)
  rc_map <- c(A = "T", C = "G", G = "C", T = "A", U = "A",
              R = "Y", Y = "R", S = "S", W = "W", K = "M", M = "K",
              B = "V", D = "H", H = "D", V = "B", N = "N", I = "N", `5` = "N")
  bases <- strsplit(seq, "", fixed = TRUE)[[1]]
  paste0(rev(unname(rc_map[bases])), collapse = "")
}

base_compatible <- function(seq_base, primer_base) {
  seq_base <- toupper(seq_base)
  primer_base <- toupper(primer_base)
  seq_chars <- strsplit(iupac_sets[[seq_base]] %||% seq_base, "", fixed = TRUE)[[1]]
  primer_chars <- strsplit(iupac_sets[[primer_base]] %||% primer_base, "", fixed = TRUE)[[1]]
  length(intersect(seq_chars, primer_chars)) > 0
}

find_best_end_overlap <- function(sequence, primer_sequence, side = c("left", "right"), min_overlap = 8L) {
  side <- match.arg(side)
  seq_chars <- strsplit(str_to_upper(sequence), "", fixed = TRUE)[[1]]
  primer_chars <- strsplit(str_to_upper(primer_sequence), "", fixed = TRUE)[[1]]
  max_overlap <- min(length(seq_chars), length(primer_chars))
  if (max_overlap < min_overlap) return(0L)

  for (overlap in seq(max_overlap, min_overlap, by = -1L)) {
    if (side == "left") {
      seq_window <- seq_chars[seq_len(overlap)]
    } else {
      seq_window <- tail(seq_chars, overlap)
    }
    primer_window <- primer_chars[seq_len(overlap)]
    matched <- vapply(seq_len(overlap), function(i) base_compatible(seq_window[[i]], primer_window[[i]]), logical(1))
    if (all(matched)) return(overlap)
  }
  0L
}

# Primer-like IUPAC symbols commonly found in degenerate primer regions
primer_like_codes <- c("R", "Y", "W", "S", "K", "M", "B", "D", "H", "V", "N")
called_codes <- c("A", "C", "G", "T")

fasta_raw <- read_fasta_simple(input_fasta)

input_seq_summary <- fasta_raw %>%
  mutate(
    seqName = header,
    sequence = str_to_upper(sequence),
    seq_chars = map(sequence, ~strsplit(.x, "", fixed = TRUE)[[1]]),
    length = map_int(seq_chars, length),
    ambiguous_count = map_int(seq_chars, ~sum(!(.x %in% called_codes))),
    ambiguous_chars = map_chr(seq_chars, ~{
      x <- sort(unique(.x[!(.x %in% called_codes)]))
      if (length(x) == 0) "" else paste(x, collapse = ",")
    }),
    primer_like_count = map_int(seq_chars, ~sum(.x %in% primer_like_codes)),
    primer_like_chars = map_chr(seq_chars, ~{
      x <- sort(unique(.x[.x %in% primer_like_codes]))
      if (length(x) == 0) "" else paste(x, collapse = ",")
    })
  ) %>%
  select(seqName, length, ambiguous_count, ambiguous_chars,
         primer_like_count, primer_like_chars, sequence)

# ── Primer catalog and end-overlap scan ──────────────────────────────────────
primer_min_overlap <- 8L
primer_catalog <- read_fasta_simple(file.path(proj_dir, "data", "Primers", "PCR_primers", "primers.fa")) %>%
  mutate(
    parsed = map(header, parse_primer_header),
    primer_name = map_chr(parsed, "name"),
    primer_direction = map_chr(parsed, ~.x$meta$direction %||% ""),
    primer_region = map_chr(parsed, ~.x$meta$region %||% ""),
    primer_comment = map_chr(parsed, ~.x$meta$comment %||% ""),
    primer_sequence = str_to_upper(sequence),
    primer_rc_sequence = map_chr(primer_sequence, primer_rc),
    primer_end = if_else(primer_direction == "F", "5' end", "3' end")
  ) %>%
  select(primer_name, primer_direction, primer_region, primer_comment, primer_sequence, primer_rc_sequence, primer_end)

primer_hits <- if (nrow(input_seq_summary) > 0 && nrow(primer_catalog) > 0) {
  map_dfr(seq_len(nrow(input_seq_summary)), function(i) {
    seq_row <- input_seq_summary[i, ]
    map_dfr(seq_len(nrow(primer_catalog)), function(j) {
      primer_row <- primer_catalog[j, ]
      scan_seq <- if (primer_row$primer_direction == "F") primer_row$primer_sequence[[1]] else primer_row$primer_rc_sequence[[1]]
      left_overlap <- find_best_end_overlap(seq_row$sequence[[1]], scan_seq, side = "left", min_overlap = primer_min_overlap)
      right_overlap <- find_best_end_overlap(seq_row$sequence[[1]], scan_seq, side = "right", min_overlap = primer_min_overlap)
      hits <- tibble(
        seqName = seq_row$seqName[[1]],
        sequence_length = seq_row$length[[1]],
        primer_name = primer_row$primer_name[[1]],
        primer_direction = primer_row$primer_direction[[1]],
        primer_region = primer_row$primer_region[[1]],
        primer_end = primer_row$primer_end[[1]],
        match_side = c("left", "right"),
        overlap_nt = c(left_overlap, right_overlap)
      ) %>%
        filter(overlap_nt >= primer_min_overlap)

      if (nrow(hits) == 0) return(tibble())

      hits %>%
        mutate(
          start = if_else(match_side == "left", 1L, sequence_length - overlap_nt + 1L),
          end = if_else(match_side == "left", overlap_nt, sequence_length),
          matched_sequence = substr(seq_row$sequence[[1]], start, end),
          matched_fraction = round(overlap_nt / nchar(scan_seq), 2),
          end_label = if_else(match_side == "left", "5' end", "3' end")
        )
    })
  })
} else {
  tibble()
}

primer_hits_table <- if (nrow(primer_hits) > 0) {
  primer_hits %>%
    arrange(seqName, match_side, desc(overlap_nt), primer_name)
} else {
  tibble()
}

primer_hit_plot <- if (nrow(primer_hits_table) > 0) {
  plot_df <- primer_hits_table %>%
    mutate(
      seqName = factor(seqName, levels = unique(seqName)),
      primer_label = if_else(primer_region == "", primer_name, paste0(primer_name, " (", primer_region, ")"))
    )

  ggplot(plot_df, aes(y = seqName)) +
    geom_segment(aes(x = start, xend = end, yend = seqName, color = primer_label),
                 linewidth = 5.5, lineend = "round", alpha = 0.9) +
    geom_point(aes(x = start), size = 1.8, color = "#111827") +
    geom_point(aes(x = end), size = 1.8, color = "#111827") +
    facet_wrap(~match_side, ncol = 1, scales = "free_y") +
    scale_x_continuous(expand = expansion(mult = c(0.02, 0.04))) +
    labs(
      title = "Potential primer remnants on input sequences",
      subtitle = paste0("Compatible end-overlaps of at least ", primer_min_overlap, " nt against built-in primer set"),
      x = "Position in input sequence (nt)",
      y = NULL,
      color = "Primer"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
} else {
  NULL
}

# ── Primer / input-sequence coverage map ────────────────────────────────────
# Coordinate system: NC_001489.1 full-genome positions.
# The 349 bp NextClade reference maps to 2919-3267 (offset = 2918 to convert
# 1-based 349 bp coords → full-genome coords).
nc_ref_offset    <- 2918L
coverage_map_win <- c(2798L, 3475L)  # display window

# Best primer binding sites (full-genome) from GFF3 analysis
primerbind <- tibble(
  primer_name = c("HAV_6.1", "HAV_8.2", "HAV_11", "HAV_10"),
  fg_start    = c(2832L, 2886L, 3375L, 3417L),
  fg_end      = c(2857L, 2914L, 3403L, 3445L),
  direction   = c("F", "F", "R", "R"),
  role        = c("First-outer", "nested-inner", "nested-inner", "First-outer")
)

# Input sequence positions in full-genome coords
seq_fg <- nc_lineages_raw %>%
  select(seqName, alignmentStart, alignmentEnd) %>%
  mutate(
    fg_start = if_else(!is.na(alignmentStart), as.integer(alignmentStart) + nc_ref_offset, NA_integer_),
    fg_end   = if_else(!is.na(alignmentEnd),   as.integer(alignmentEnd)   + nc_ref_offset, NA_integer_),
    aligned  = !is.na(fg_start)
  )

# Which sequence ends have flagged primer remnants?
left_hit_seqs  <- if (nrow(primer_hits_table) > 0) {
  primer_hits_table %>% filter(match_side == "left")  %>% pull(seqName) %>% unique()
} else {
  character()
}
right_hit_seqs <- if (nrow(primer_hits_table) > 0) {
  primer_hits_table %>% filter(match_side == "right") %>% pull(seqName) %>% unique()
} else {
  character()
}

# Build y positions: reference at bottom, then primers, then input sequences
bar_h     <- 0.37
prim_gap  <- 1.4
ref_y     <- 0
n_prim    <- nrow(primerbind)
prim_ys   <- setNames(seq_len(n_prim) * 1.0 + 0.5, rev(primerbind$primer_name))
seq_names_ord <- rev(seq_fg$seqName)
n_seqs    <- length(seq_names_ord)
seq_ys    <- setNames(seq_along(seq_names_ord) + n_prim + prim_gap, seq_names_ord)

prim_colors <- c(
  "HAV_6.1" = "#1f77b4", "HAV_8.2" = "#2ca02c",
  "HAV_11"  = "#d62728", "HAV_10"  = "#ff7f0e"
)

ref_df   <- tibble(xmin = coverage_map_win[[1]], xmax = coverage_map_win[[2]],
                   ymin = ref_y - bar_h, ymax = ref_y + bar_h)

primerbind_df <- primerbind %>%
  mutate(
    y    = prim_ys[primer_name],
    ymin = y - bar_h, ymax = y + bar_h,
    arw_x    = if_else(direction == "F", fg_end,   fg_start),
    arw_xend = if_else(direction == "F", fg_end + 16L, fg_start - 16L)
  )

seq_df <- seq_fg %>%
  mutate(
    y    = seq_ys[seqName],
    ymin = y - bar_h, ymax = y + bar_h,
    xmin = if_else(aligned, fg_start, coverage_map_win[[1]]),
    xmax = if_else(aligned, fg_end,   coverage_map_win[[2]]),
    has_left_hit  = seqName %in% left_hit_seqs,
    has_right_hit = seqName %in% right_hit_seqs
  )

# Primer-remnant highlight bars at sequence ends (orange overlay)
hit_left_df <- seq_df %>% filter(aligned, has_left_hit) %>%
  mutate(xmin_h = fg_start, xmax_h = fg_start + 29L)
hit_right_df <- seq_df %>% filter(aligned, has_right_hit) %>%
  mutate(xmin_h = fg_end - 29L, xmax_h = fg_end)
hit_overlay_df <- bind_rows(hit_left_df, hit_right_df)

coverage_map_plot <- ggplot() +
  # Product region shading
  annotate("rect", xmin = 2832, xmax = 3445,
           ymin = -Inf, ymax = Inf, fill = "#bbbbbb", alpha = 0.15) +
  annotate("rect", xmin = 2886, xmax = 3403,
           ymin = -Inf, ymax = Inf, fill = "#888888", alpha = 0.15) +
  # Reference bar
  geom_rect(data = ref_df,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#444444", color = NA, inherit.aes = FALSE) +
  # Aligned input sequence bars
  geom_rect(data = seq_df %>% filter(aligned),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#aec6e8", color = "#1a6bb5", linewidth = 0.3, inherit.aes = FALSE) +
  # Unaligned sequences (dashed)
  geom_rect(data = seq_df %>% filter(!aligned),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#e8e8e8", color = "#999999", linewidth = 0.3,
            linetype = "dashed", inherit.aes = FALSE) +
  # Primer-remnant overlays at sequence ends
  geom_rect(data = hit_overlay_df,
            aes(xmin = xmin_h, xmax = xmax_h, ymin = ymin, ymax = ymax),
            fill = "#e05c00", alpha = 0.6, color = NA, inherit.aes = FALSE) +
  # Primer binding site bars
  geom_rect(data = primerbind_df,
            aes(xmin = fg_start, xmax = fg_end, ymin = ymin, ymax = ymax,
                fill = primer_name),
            color = "white", linewidth = 0.5, inherit.aes = FALSE) +
  scale_fill_manual(values = prim_colors, name = "Primer") +
  # Direction arrows
  geom_segment(data = primerbind_df,
               aes(x = arw_x, xend = arw_xend, y = y, yend = y,
                   color = primer_name),
               arrow = arrow(length = unit(0.12, "cm"), type = "closed"),
               linewidth = 0.9, inherit.aes = FALSE) +
  scale_color_manual(values = prim_colors, guide = "none") +
  # Axes and labels
  scale_x_continuous(
    limits = coverage_map_win,
    labels = scales::comma,
    expand = expansion(mult = 0),
    name = "NC_001489.1 position (bp)"
  ) +
  scale_y_continuous(
    breaks = c(ref_y, unname(prim_ys), unname(seq_ys)),
    labels = c("Reference", names(prim_ys), names(seq_ys)),
    expand = expansion(add = 0.9)
  ) +
  labs(
    title    = "Primer binding sites and input sequence coverage",
    subtitle = paste0(
      "Shading: outer PCR product (light, ", 2832, "-", 3445, " bp) and ",
      "inner/nested product (dark, ", 2886, "-", 3403, " bp).\n",
      "Orange = detected primer remnant at sequence end. Dashed = failed alignment."
    ),
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 9),
    legend.position    = "right"
  )

# ── Junction-focused alignment view (from Nextclade) ───────────────────────
junction_aln_path <- file.path(out_dir, "lineages", "nextclade.aligned.fasta")
junction_aln <- safe_read_dna_fasta(junction_aln_path)

junction_is_usable <- !is.null(junction_aln) && nrow(junction_aln) >= 2

junction_msa_plot <- NULL
junction_var_plot <- NULL

if (junction_is_usable) {
  junction_ids <- rownames(junction_aln)

  junction_order <- summary_tbl %>%
    select(seqName, clade, lineage_phylo, qc_status) %>%
    mutate(
      clade_rank = dplyr::coalesce(clade, "ZZZ"),
      lineage_rank = dplyr::coalesce(lineage_phylo, "ZZZ"),
      qc_rank = dplyr::coalesce(qc_status, "ZZZ")
    ) %>%
    arrange(clade_rank, lineage_rank, qc_rank, seqName) %>%
    pull(seqName)

  junction_order <- unique(c(junction_order[junction_order %in% junction_ids], junction_ids))

  junction_labels <- summary_tbl %>%
    select(seqName, clade, lineage_phylo) %>%
    mutate(
      clade = if_else(is.na(clade) | clade == "", "?", clade),
      lineage_phylo = if_else(is.na(lineage_phylo) | lineage_phylo == "", "?", lineage_phylo),
      label = paste0(seqName, " [", clade, "/", lineage_phylo, "]")
    ) %>%
    select(seqName, label)

  label_map <- setNames(junction_labels$label, junction_labels$seqName)

  dna_ss_junc <- tryCatch(Biostrings::readDNAStringSet(junction_aln_path), error = function(e) NULL)
  if (!is.null(dna_ss_junc)) {
    keep_ids <- junction_order[junction_order %in% names(dna_ss_junc)]
    if (length(keep_ids) >= 2) {
      dna_ss_junc <- dna_ss_junc[keep_ids]
      names(dna_ss_junc) <- dplyr::coalesce(label_map[names(dna_ss_junc)], names(dna_ss_junc))

      junction_msa_plot <- ggmsa::ggmsa(
        dna_ss_junc,
        start = 1,
        end = Biostrings::width(dna_ss_junc)[1],
        color = "Chemistry_NT",
        seq_name = TRUE,
        font = NULL,
        show.legend = TRUE
      ) +
        labs(
          title = "Nextclade junction alignment (full displayed window)",
          subtitle = "All batch query sequences aligned to the short VP1-2B junction reference",
          x = "Junction alignment position",
          y = NULL
        ) +
        theme_bw(base_size = 9) +
        theme(
          plot.title = element_text(face = "bold", size = 10),
          axis.text.y = element_text(size = 6.5),
          axis.text.x = element_text(size = 7)
        )
    } # end junction MSA plot
  } # end junction alignment visualization

  mat_j <- tolower(junction_aln[junction_order, , drop = FALSE])
  called_j <- c("a", "c", "g", "t")

  .cons_j <- function(col) {
    vals <- col[col %in% called_j]
    if (length(vals) == 0) return(NA_character_)
    tab <- sort(table(vals), decreasing = TRUE)
    names(tab)[1]
  }

  consensus_j <- apply(mat_j, 2, .cons_j)
  snp_count_j <- sapply(seq_len(ncol(mat_j)), function(k) {
    col <- mat_j[, k]
    cons <- consensus_j[k]
    if (is.na(cons)) return(0L)
    sum(col %in% called_j & col != cons)
  })
  var_cols_j <- which(snp_count_j > 0)

  if (length(var_cols_j) > 0) {
    mat_var_j <- mat_j[, var_cols_j, drop = FALSE]
    pos_var_j <- var_cols_j
    row_ids_j <- rownames(mat_var_j)

    display_levels_j <- rev(dplyr::coalesce(label_map[row_ids_j], row_ids_j))

    var_grid_j <- expand.grid(
      seq_id = row_ids_j,
      col_i = seq_len(ncol(mat_var_j)),
      KEEP.OUT.ATTRS = FALSE,
      stringsAsFactors = FALSE
    ) %>%
      as_tibble() %>%
      mutate(
        base = mat_var_j[cbind(match(seq_id, row_ids_j), col_i)],
        position = pos_var_j[col_i],
        seq_label = dplyr::coalesce(label_map[seq_id], seq_id),
        base_class = case_when(
          base %in% called_j ~ str_to_upper(base),
          base == "-" ~ "Gap",
          TRUE ~ "Ambig"
        )
      )

    nt_pal_j <- c(
      "A" = "#4daf4a", "C" = "#377eb8", "G" = "#ff7f00", "T" = "#e41a1c",
      "Gap" = "#f5f5f5", "Ambig" = "#6a3d9a"
    )

    junction_var_plot <- ggplot(
      var_grid_j,
      aes(x = position, y = factor(seq_label, levels = display_levels_j), fill = base_class)
    ) +
      geom_tile(color = "white", linewidth = 0.15) +
      scale_fill_manual(values = nt_pal_j, drop = FALSE) +
      labs(
        title = paste0("Nextclade junction variant-only view (", length(var_cols_j), " variable sites)"),
        x = "Junction alignment position (variable sites)",
        y = NULL,
        fill = "Base"
      ) +
      theme_bw(base_size = 10) +
      theme(
        axis.text.y = element_text(size = 7),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 10),
        legend.position = "bottom"
      )
  } # end junction variant plot
} # end junction alignment check

input_overview <- if (nrow(input_seq_summary) > 0) {
  lens <- input_seq_summary$length
  tibble(
    metric = c(
      "Total sequences",
      "Unique headers",
      "Duplicated headers",
      "Min length",
      "Max length",
      "Mean length",
      "Median length",
      "Sequences with ambiguous bases",
      "Total ambiguous bases",
      "Sequences with primer-like IUPAC codes",
      "Total primer-like IUPAC bases"
    ),
    value = c(
      nrow(input_seq_summary),
      n_distinct(input_seq_summary$seqName),
      nrow(input_seq_summary) - n_distinct(input_seq_summary$seqName),
      min(lens),
      max(lens),
      round(mean(lens), 2),
      round(stats::median(lens), 2),
      sum(input_seq_summary$ambiguous_count > 0),
      sum(input_seq_summary$ambiguous_count),
      sum(input_seq_summary$primer_like_count > 0),
      sum(input_seq_summary$primer_like_count)
    )
  )
} else {
  tibble(metric = "Input FASTA", value = "Not found")
}

# Ensure lineage_phylo column exists
if (!"lineage_phylo" %in% names(summary_tbl)) summary_tbl$lineage_phylo <- NA_character_

# Is lineage assigned (not a NODE_ placeholder)?
is_lineage <- function(x) !is.na(x) & !str_starts(x, "NODE_")

summary_tbl <- summary_tbl %>%
  mutate(
    clade_agree_blast = case_when(
      !is.na(blast_genotype) & !is.na(clade) ~
        str_to_upper(replace_na(blast_genotype, "")) ==
          str_to_upper(replace_na(clade, "")),
      TRUE ~ NA
    )
  )

# ── Read samplesheet for LabWareID and SequenceID ─────────────────────────────
.ss_sample_info <- tibble()
.ss_path <- file.path(batch_dir, "auto_samplesheet.tsv")
if (!file.exists(.ss_path)) {
  # Try finding any samplesheet
  .ss_files <- list.files(batch_dir, pattern = "samplesheet\\.tsv$", full.names = TRUE)
  if (length(.ss_files) > 0) .ss_path <- .ss_files[1]
}
if (file.exists(.ss_path)) {
  .ss_sample_info <- safe_tsv(.ss_path) %>%
    select(Sample, any_of(c("LabWareID", "SekvensID"))) %>%
    distinct()
}

# ── Add input sequence length to summary ─────────────────────────────────────
.seq_lengths <- if (nrow(input_seq_summary) > 0) {
  input_seq_summary %>%
    transmute(seqName, input_length = length)
} else {
  tibble(seqName = character(), input_length = integer())
}

# ── Count BLAST database occurrences per variant and find similar variants ────
# This helps identify how common the detected variant is in the reference database
# and warns if other variants are equally similar (SNP distance <= 3)
.db_variant_summary <- {
  metadata_tsv_df <- safe_tsv(metadata_tsv)


  if (nrow(metadata_tsv_df) > 0 && "variant" %in% names(metadata_tsv_df) && nrow(blast_raw) > 0) {
    # Normalize metadata variant into one canonical column for counting/joining.
    metadata_variant <- metadata_tsv_df %>%
      transmute(
        id,
        variant = as.character(.data[["variant"]])
      ) %>%
      filter(!is.na(variant), variant != "")

    # Count sequences per variant in the local database.
    variant_counts <- metadata_variant %>%
      count(variant, name = "db_count") %>%
      rename(blast_variant = variant)

    # Join BLAST results with metadata to get variant per hit.
    blast_with_variant <- blast_raw %>%
      left_join(metadata_variant %>% rename(sseqid = id), by = "sseqid")

    # For each sample/query, find alternative variant within <=3 SNP total
    # from the query sequence (absolute threshold, not relative to best hit).
    similar_variants <- blast_with_variant %>%
      filter(!is.na(variant)) %>%
      group_by(qseqid) %>%
      arrange(mismatch, desc(bitscore), .by_group = TRUE) %>%
      mutate(
        best_sseqid = first(sseqid),
        best_variant = first(variant)
      ) %>%
      filter(
        sseqid != best_sseqid,
        variant != best_variant,
        mismatch <= 3
      ) %>%
      summarise(
        sseqid = first(best_sseqid),
        similar_variants = paste(unique(sort(variant)), collapse = ", "),
        .groups = "drop"
      )

    list(
      counts = variant_counts,
      similar = similar_variants
    )
  } else {
    list(
      counts = tibble(blast_variant = character(), db_count = integer()),
      similar = tibble(qseqid = character(), sseqid = character(), similar_variants = character())
    )
  }
}

.blast_variant_counts <- .db_variant_summary$counts %>%
  rename(db_variant_count = db_count)

.similar_variants_map <- .db_variant_summary$similar %>%
  setNames(c("qseqid", "sseqid", "similar_variants")) %>%
  {if (nrow(.) > 0) . else tibble(sseqid = character(), similar_variants = character())}

# ── Batch triage: one confidence flag per sample ──────────────────────────────
# Lets the expert skip the obvious calls and focus on the few that need a look.
# The flag is driven by genotype confidence (the report's primary question), not
# by raw BLAST %identity — for WGS the nearest reference rarely reaches
# outbreak-level identity, so a %id threshold would flag everything.
.called <- if (nrow(input_seq_summary) > 0) {
  input_seq_summary %>%
    transmute(join_key   = word(seqName, 1),
              called_frac = if_else(length > 0, (length - ambiguous_count) / length, NA_real_))
} else tibble(join_key = character(), called_frac = numeric())

triage_tbl <- summary_tbl %>%
  mutate(join_key = word(seqName, 1)) %>%
  left_join(.called, by = "join_key") %>%
  left_join(.seq_lengths, by = "seqName") %>%
  left_join(.ss_sample_info, by = c("seqName" = "Sample")) %>%
  left_join(.blast_variant_counts, by = "blast_variant") %>%
  left_join(.db_variant_summary$similar, 
            by = c("qseqid" = "qseqid", "sseqid" = "sseqid")) %>%
  mutate(
    # Format similar variants warning
    similar_variants_msg = map_chr(similar_variants, ~ {
      if (is.na(.x) || .x == "") return("")
      ids <- str_split(.x, ",\\s*", simplify = FALSE)[[1]]
      ids <- trimws(ids)
      lines <- split(ids, ceiling(seq_along(ids) / 2))
      wrapped_ids <- map_chr(lines, ~ paste(.x, collapse = ", "))
      paste0("⚠ Also similar (≤3 SNPs):<br>",
             paste(wrapped_ids, collapse = "<br>"),
             "<br>— see heatmap")
    }),
    flag = case_when(
      is.na(clade)                                   ~ "No call",
      !is.na(clade_agree_blast) & !clade_agree_blast ~ "Review",
      TRUE                                           ~ "Clear"
    ),
    flag_reason = case_when(
      flag == "No call" & !is.na(called_frac) ~ sprintf("No NextClade call (%.0f%% of genome called)", 100 * called_frac),
      flag == "No call"                       ~ "No NextClade call",
      flag == "Review"                        ~ "NextClade and BLAST genotype disagree",
      TRUE                                    ~ ""
    )
  )
```

# `r batch_name` — HAV Analysis Summary

This report compares complementary HAV analyses run on **`r batch_name`**:

| Analysis | Description |
|---|---|
| **NC-lineages** | NextClade with `hav-vp1-2b-lineages` dataset — assigns clade (IA/IB/IIA/IIIA) + sub-genotype `lineage_phylo` |
| **BLAST** | BLAST against local Norwegian HAV sequence database |

---

# Batch overview {.tabset}

## Verdict table

```{r triage-table}
.flag_col <- c(Clear = "#2ca02c", Review = "#d62728", `No call` = "#7f7f7f")
.flag_ord <- c("Review", "No call", "Clear")

triage_tbl %>%
  arrange(factor(flag, levels = .flag_ord), seqName) %>%
  transmute(
    `LabWareID`     = if_else(!is.na(LabWareID), LabWareID, "—"),
    `SequenceID`    = if_else(!is.na(SekvensID), SekvensID, "—"),
    `Genotype (NC)` = if_else(!is.na(clade), clade, "—"),
    `Genotype (BLAST)` = if_else(!is.na(blast_genotype), blast_genotype, "—"),
    `Agree`         = case_when(is.na(clade_agree_blast) ~ "—",
                                clade_agree_blast ~ "✓", TRUE ~ "✗"),
    `BLAST top hit` = if_else(!is.na(sseqid), sseqid, "—"),
    `BLAST variant` = if_else(!is.na(blast_variant), blast_variant, "—"),
    `BLAST SNPs`    = if_else(!is.na(blast_mismatch), as.character(blast_mismatch), "—"),
    `DB variant count` = if_else(!is.na(db_variant_count), as.character(db_variant_count), "—"),
    `Similar BLAST variant` = if_else(similar_variants_msg != "", similar_variants_msg, "—"),
    `Input length`  = if_else(!is.na(input_length), as.character(input_length), "—"),
    `Called %`      = if_else(!is.na(called_frac), percent(called_frac, 1), "—"),
    `Flag`          = cell_spec(flag, color = "white", background = .flag_col[flag], bold = TRUE),
    `Notes`         = flag_reason
  ) -> .triage_display

# Calculate background colors for BLAST SNPs column (same as Per-sequence summary)
.snp_colors <- .triage_display$`BLAST SNPs` %>%
  map_chr(~ {
    if (. == "—") {
      ""  # Empty string for no background
    } else {
      snp_count <- as.numeric(.)
      if (snp_count == 0) {
        "#c8e6c9"  # Light green for 0 SNPs
      } else if (snp_count <= 2) {
        "#fff9c4"  # Light yellow for 1-2 SNPs
      } else {
        ""  # Empty string for no background
      }
    }
  })

# Calculate background colors for Input length column (yellow if != 460)
.input_length_colors <- .triage_display$`Input length` %>%
  map_chr(~ {
    if (. == "—") {
      ""
    } else {
      length_val <- as.numeric(.)
      if (!is.na(length_val) && length_val != 460) {
        "rgb(230, 218, 111)"  # Light yellow for non-460 lengths
      } else {
        ""
      }
    }
  })

.triage_display %>%
  kbl(escape = FALSE,
      table.attr = 'style="width: auto;"',
      caption = "Batch overview — sorted with items needing review first. Green = clear, red = genotype disagreement, grey = no NextClade call (usually low coverage). Called % = percentage of sequence bases that are unambiguous (A/C/G/T; excludes N and other ambiguous codes). DB variant count = number of sequences in reference database with same BLAST variant.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(1, width = "10em", extra_css = "white-space: nowrap;") %>%  # LabWareID
  column_spec(2, width = "12em", extra_css = "white-space: nowrap;") %>%  # SequenceID
  column_spec(3, width = "10em", extra_css = "white-space: nowrap;") %>%  # Genotype (NC)
  column_spec(4, width = "12em", extra_css = "white-space: nowrap;") %>%  # Genotype (BLAST)
  column_spec(6, width = "14em", extra_css = "white-space: nowrap;") %>%  # BLAST top hit
  column_spec(7, width = "16em", extra_css = "white-space: nowrap;") %>%  # BLAST variant
  column_spec(8, background = .snp_colors, bold = TRUE) %>%  # Color code BLAST SNPs
  column_spec(10, width = "20em", extra_css = "min-width: 18em; white-space: normal; line-height: 1.2; word-break: break-word;") %>%  # Similar BLAST variant
  column_spec(11, background = .input_length_colors) %>%  # Color code Input length
  scroll_box(width = "100%")
```

`r {n_clear <- sum(triage_tbl$flag=="Clear"); n_rev <- sum(triage_tbl$flag=="Review"); n_no <- sum(triage_tbl$flag=="No call"); sprintf("**%d clear · %d to review · %d no call** of %d samples.", n_clear, n_rev, n_no, nrow(triage_tbl))}`

## Placement tree (NextClade)

```{r placement-tree, fig.width=11, fig.height=11}
.nwk_path <- file.path(out_dir, "lineages", "nextclade.nwk")
.ptr <- if (file.exists(.nwk_path)) tryCatch(read.tree(.nwk_path), error = function(e) NULL) else NULL

if (is.null(.ptr)) {
  cat("NextClade placement tree (`nextclade.nwk`) not found — run the NextClade step first.")
} else {
  .qids <- word(all_queries, 1)
  .tipmeta <- tibble(label = .ptr$tip.label) %>%
    left_join(community_meta %>% transmute(label = id, ref_clade = community_clade), by = "label") %>%
    left_join(triage_tbl %>% transmute(label = word(seqName, 1), q_clade = clade), by = "label") %>%
    mutate(
      is_query = label %in% .qids,
      Genotype = coalesce(q_clade, ref_clade),
      Genotype = if_else(is.na(Genotype) | Genotype %in% c("", "NA"), "unassigned", str_to_upper(Genotype))
    )

  .gtypes <- sort(setdiff(unique(.tipmeta$Genotype), "unassigned"))
  .pal <- setNames(scales::hue_pal()(length(.gtypes)), .gtypes)
  .pal["unassigned"] <- "grey80"

  p_place <- ggtree(.ptr, layout = "fan", open.angle = 12, size = 0.15, color = "grey55") %<+% .tipmeta +
    geom_tippoint(aes(color = Genotype, size = is_query, alpha = is_query)) +
    scale_size_manual(values = c(`FALSE` = 0.5, `TRUE` = 3.2), guide = "none") +
    scale_alpha_manual(values = c(`FALSE` = 0.55, `TRUE` = 1), guide = "none") +
    scale_color_manual(values = .pal, name = "Genotype / clade",
                       breaks = .gtypes) +
    ggtree::geom_tiplab2(aes(label = ifelse(is_query, label, NA), color = Genotype),
                         size = 3, fontface = "bold", offset = 0.001, na.rm = TRUE) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    labs(title = "Batch samples placed on the NextClade reference tree",
         subtitle = sprintf("%d reference tips + %d batch samples (bold red-rim labels). Placement uses the VP1-2B junction.",
                            sum(!.tipmeta$is_query), sum(.tipmeta$is_query))) +
    theme(legend.position = "right",
          plot.title = element_text(face = "bold"))

  print(p_place)
}
```

---

# 0. Input sequence summary

## Per-sequence input QC

```{r input-per-seq}
if (nrow(input_seq_summary) > 0) {
  input_seq_summary %>%
    transmute(
      Sequence = seqName,
      Length = length,
      Ambiguous = ambiguous_count
    ) %>%
    kbl() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("Input FASTA not found:", input_fasta)
}
```

---

```{r amplicon-coverage, results='asis', fig.width=12}
if (has_amplicon && nrow(amplicon_long) > 0) {
  cat("# Amplicon coverage (whole-genome / ViroConstrictor) {.tabset}\n\n")

  # ── Tab 1: coverage barplot, one bar per amplicon, faceted per sample ───────
  cat("## Coverage barplot\n\n")

  .mincov <- if (!is.na(mincov_val)) mincov_val else 20

  # Per-sample genotype (NextClade clade) for the facet headers + grid ordering.
  .geno_lookup <- triage_tbl %>%
    transmute(sample = word(seqName, 1),
              geno   = harmonise_genotype(clade)) %>%
    distinct(sample, .keep_all = TRUE)

  amp_plot <- amplicon_long %>%
    left_join(.geno_lookup, by = "sample") %>%
    mutate(
      geno     = if_else(is.na(geno), "—", geno),
      passes   = coverage >= .mincov,
      cov_plot = pmax(coverage, 0.5),       # floor so zeros are visible on log scale
      facet_label = paste0(sample, " · ", geno)
    )

  # Order facets by genotype (no-call last), then sample, so same-genotype
  # samples sit together in the grid.
  .facet_levels <- amp_plot %>%
    distinct(sample, geno, facet_label) %>%
    arrange(geno == "—", geno, sample) %>%
    pull(facet_label)
  amp_plot <- amp_plot %>%
    mutate(facet_label = factor(facet_label, levels = .facet_levels))

  n_samples <- dplyr::n_distinct(amp_plot$sample)
  p_amp <- ggplot(amp_plot, aes(x = factor(amplicon), y = cov_plot, fill = passes)) +
    geom_col() +
    geom_hline(yintercept = .mincov, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
    facet_wrap(~ facet_label, ncol = 4) +
    scale_y_log10(labels = scales::label_number()) +
    scale_fill_manual(
      values = c(`TRUE` = "#2ca02c", `FALSE` = "#d62728"),
      labels = c(`TRUE` = paste0("≥ MinCov (", .mincov, "×)"),
                 `FALSE` = paste0("< MinCov (", .mincov, "×)")),
      name = NULL, drop = FALSE
    ) +
    labs(
      x = "Amplicon", y = "Mean coverage (×, log scale)",
      title = "Per-amplicon mean coverage by sample",
      subtitle = paste0("Dashed line = MinCov threshold (", .mincov,
                        "×). Bars below it (red) failed the consensus depth filter.")
    ) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom",
          panel.grid.major.x = element_blank())

  .tmp_amp <- tempfile(fileext = ".png")
  ggsave(.tmp_amp, plot = p_amp, width = 12,
         height = 2.4 * ceiling(n_samples / 4) + 1, dpi = 150, limitsize = FALSE)
  cat(sprintf('<img src="%s" style="width:100%%;"/>\n\n', knitr::image_uri(.tmp_amp)))
  file.remove(.tmp_amp)

  # Mean coverage per amplicon across the batch (compact summary table)
  amp_mean <- amplicon_long %>%
    group_by(Amplicon = amplicon) %>%
    summarise(
      `Mean coverage (×)`   = round(mean(coverage, na.rm = TRUE), 1),
      `Median coverage (×)` = round(median(coverage, na.rm = TRUE), 1),
      `Samples ≥ MinCov`    = sum(coverage >= .mincov, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(`Samples ≥ MinCov` = paste0(`Samples ≥ MinCov`, " / ", n_samples))
  cat(kbl(amp_mean, caption = "Per-amplicon coverage across the batch") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width = FALSE) %>%
        as.character())
  cat("\n\n")

  # ── Tab 2: primer scheme table (primer name ↔ amplicon number) ──────────────
  cat("## Primer scheme\n\n")
  if (nrow(primer_scheme_tbl) > 0) {
    .ptbl <- primer_scheme_tbl %>%
      select(Scheme = scheme, Amplicon = amplicon,
             `Left primer`            = any_of("primer_LEFT"),
             `Left primer (original)` = any_of("original_LEFT"),
             `Left sequence`          = any_of("sequence_LEFT"),
             `Right primer`            = any_of("primer_RIGHT"),
             `Right primer (original)` = any_of("original_RIGHT"),
             `Right sequence`          = any_of("sequence_RIGHT"))
    cat(kbl(.ptbl, caption = "Primer scheme: primer names (ViroConstrictor & original) and sequences per amplicon") %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                        full_width = FALSE) %>%
          as.character())
    cat("\n\n")
  } else {
    cat("*Primer scheme FASTA referenced by the samplesheet could not be read.*\n\n")
  }

  cat("---\n\n")
}
```

# 1. Per-sequence summary

**NextClade (NC)** designates lineage assignments from the `hav-vp1-2b-lineages` dataset.

```{r summary-table}
clade_comp <- summary_tbl %>%
  select(seqName, clade, lineage_phylo, blast_genotype, blast_variant) %>%
  mutate(
    blast_vs_clade = case_when(
      is.na(blast_genotype) | is.na(clade) ~ "No BLAST hit",
      str_detect(
        str_to_upper(clade),
        str_remove(str_to_upper(replace_na(blast_genotype, "")), "[Aa]$|[Bb]$")
      ) ~ "Agreement",
      TRUE ~ "DISCORDANT"
    )
  )

summary_tbl %>%
  mutate(
    coverage       = if_else(!is.na(coverage) & is.numeric(coverage), percent(coverage, 0.1), "—"),
    blast_pident   = if_else(!is.na(blast_pident),   paste0(round(blast_pident, 1), "%"), "—"),
    blast_mismatch = if_else(!is.na(blast_mismatch), as.character(blast_mismatch),        "—"),
    blast_qcovs    = if_else(!is.na(blast_qcovs),    paste0(blast_qcovs, "%"),           "—"),
    clade          = if_else(!is.na(clade),           clade,                               "—"),
    lineage_phylo  = if_else(!is.na(lineage_phylo),   lineage_phylo,                       "—"),
    blast_variant  = if_else(!is.na(blast_variant),   blast_variant,                       "—"),
    blast_genotype = if_else(!is.na(blast_genotype),  blast_genotype,                      "—")
  ) %>%
  left_join(clade_comp %>% select(seqName, blast_vs_clade), by = "seqName") %>%
  select(
    Sequence          = seqName,
    `NC Clade`        = clade,
    `Lineage (phylo)` = lineage_phylo,
    `NC QC`           = qc_status,
    `BLAST genotype`  = blast_genotype,
    `BLAST top hit`   = sseqid,
    `BLAST %id`       = blast_pident,
    `BLAST SNPs`      = blast_mismatch,
    `BLAST qcov%`     = blast_qcovs,
    `BLAST variant`   = blast_variant
  ) -> summary_display

# Calculate background colors for BLAST SNPs column
snp_colors <- summary_display$`BLAST SNPs` %>%
  map_chr(~ {
    if (. == "—") {
      ""  # Empty string for no background
    } else {
      snp_count <- as.numeric(.)
      if (snp_count == 0) {
        "#c8e6c9"  # Light green for 0 SNPs
      } else if (snp_count <= 2) {
        "#fff9c4"  # Light yellow for 1-2 SNPs
      } else {
        ""  # Empty string for no background
      }
    }
  })

summary_display %>%
  kbl(escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(3, bold = TRUE, color = "#1a6bb5") %>%
  column_spec(5, bold = TRUE) %>%
  column_spec(8, bold = TRUE, background = snp_colors)
```

---

# 2. BLAST analysis detail {.tabset}

## Top hits per sequence

```{r blast-detail}
if (nrow(blast_raw) > 0) {
  # Prepare metadata join with safe column selection
  meta_to_join <- meta %>%
    select(id, any_of("date")) %>%
    rename(sseqid = id)
  
  if ("date" %in% names(meta_to_join)) {
    meta_to_join <- meta_to_join %>% rename(hit_date = date)
  }
  
  blast_top %>%
    filter(hit_rank <= 10) %>%
    left_join(meta_to_join, by = "sseqid") %>%
    select(
      Query    = qseqid,
      Rank     = hit_rank,
      Hit      = sseqid,
      `% identity` = pident,
      `Query cov%`  = qcovs,
      `SNPs`        = mismatch,
      `Bitscore`    = bitscore,
      any_of(c("hit_genotype", "hit_variant", "hit_date"))
    ) %>%
    rename_with(~case_when(
      . == "hit_genotype" ~ "Genotype",
      . == "hit_variant" ~ "Variant",
      . == "hit_date" ~ "Date",
      TRUE ~ .
    )) %>%
    mutate(
      `% identity` = round(`% identity`, 2),
      across(any_of(c("Genotype", "Variant", "Date")), 
             ~replace_na(as.character(.), "—"))
    ) %>%
    kbl() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    scroll_box(height = "500px")
} else {
  cat("No BLAST results found.")
}
```



---

# 4. Per-sequence details {.tabset}

<div id="tree-lightbox" onclick="this.style.display='none'" style="display:none;position:fixed;top:0;left:0;width:100%;height:100%;background:rgba(0,0,0,.92);z-index:9999;cursor:zoom-out;overflow:auto;display:none;align-items:center;justify-content:center;">
  <img id="tree-lightbox-img" style="max-width:95vw;max-height:95vh;display:block;margin:auto;"/>
  <div style="position:fixed;top:12px;right:20px;color:#fff;font-size:28px;cursor:pointer;" onclick="document.getElementById('tree-lightbox').style.display='none'">&times;</div>
</div>

## {.unlisted .unnumbered}

```{r per-seq-details, results='asis'}
suppressPackageStartupMessages({
  library(pheatmap)
  library(ggmsa)
})

# ── Shared setup ──────────────────────────────────────────────────────────────
ref_meta <- meta %>%
  select(id, genotype, variant) %>%
  mutate(
    id_norm = norm_id(id),
    genotype = str_to_upper(genotype),
    variant = if_else(
      is.na(variant) | variant == "",
      NA_character_,
      variant
    )
  )

query_lp <- nc_lineages %>%
  select(seqName, lineage_phylo) %>%
  filter(!is.na(lineage_phylo), is_lineage(lineage_phylo))

query_ids <- summary_tbl$seqName
called    <- c("a", "c", "g", "t")

# Normalize query IDs to match tree directory names (first token only, trimmed)
query_ids_normalized <- word(query_ids, 1)

.is_called <- function(x) {
  x %in% called
}

.consensus_base <- function(col) {
  vals <- col[.is_called(col)]
  if (length(vals) == 0) return(NA_character_)
  tab <- sort(table(vals), decreasing = TRUE)
  names(tab)[1]
}

.col_entropy <- function(col) {
  vals <- col[.is_called(col)]
  if (length(vals) == 0) return(NA_real_)
  p <- as.numeric(table(vals)) / length(vals)
  -sum(p * log2(p))
}

.roll_mean <- function(x, k = 15) {
  n <- length(x)
  if (n == 0) return(numeric(0))
  k <- max(1L, as.integer(k))
  h <- floor(k / 2)
  sapply(seq_len(n), function(i) {
    lo <- max(1L, i - h)
    hi <- min(n, i + h)
    mean(x[lo:hi], na.rm = TRUE)
  })
}

.rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || diff(rng) == 0) return(rep(0, length(x)))
  (x - rng[1]) / diff(rng)
}

.pick_zoom_windows <- function(score, n_windows = 5, width = 35) {
  if (length(score) == 0 || all(!is.finite(score))) {
    return(tibble(start = integer(), end = integer(), center = integer(), score = double()))
  }
  width <- max(5L, as.integer(width))
  half  <- floor(width / 2)
  ord   <- order(score, decreasing = TRUE, na.last = NA)
  chosen <- integer()
  for (idx in ord) {
    if (length(chosen) >= n_windows) break
    if (length(chosen) == 0 || all(abs(idx - chosen) > width)) {
      chosen <- c(chosen, idx)
    }
  }
  if (length(chosen) == 0) {
    return(tibble(start = integer(), end = integer(), center = integer(), score = double()))
  }
  tibble(
    center = chosen,
    start  = pmax(1L, chosen - half),
    end    = pmin(length(score), chosen + half),
    score  = score[chosen]
  ) %>%
    arrange(desc(score))
}

.embed_img <- function(path) {
  cat(sprintf(
    '<img src="%s" onclick="document.getElementById(\'tree-lightbox-img\').src=this.src; var lb=document.getElementById(\'tree-lightbox\'); lb.style.display=\'flex\';" style="width:100%%; cursor:zoom-in; border:1px solid #ddd; border-radius:4px; margin-bottom:12px;" title="Click to enlarge"/>\n',
    knitr::image_uri(path)
  ))
}

treefile_exists <- sapply(query_ids_normalized, function(id) file.exists(file.path(trees_dir, id, "tree.treefile")))
aln_exists <- sapply(query_ids_normalized, function(id) {
  # Prefer BLAST-only alignment if available, otherwise fall back to full alignment
  blast_only <- file.exists(file.path(trees_dir, id, "aligned_trimmed_blast_only.fa"))
  full_aln <- file.exists(file.path(trees_dir, id, paste0(id, ".aligned_trimmed.fasta")))
  blast_only || full_aln
})

if (length(query_ids) == 0) {
  cat("## Status\n\n")
  cat("*No sequences available in summary table for per-sequence details.*\n\n")
} else if (!any(treefile_exists) && !any(aln_exists)) {
  cat("## Status\n\n")
  cat("Per-sequence tree outputs are not present for this batch, so detailed tree/alignment panels are skipped.\n\n")

  status_tbl <- tibble(
    Sequence = query_ids,
    `Tree file` = if_else(treefile_exists, "Found", "Missing"),
    `Tree alignment` = if_else(aln_exists, "Found", "Missing")
  )

  cat(kbl(status_tbl, caption = "Per-sequence tree artifacts") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width = TRUE) %>%
        as.character())
  cat("\n\n")

  cat("Run this first (from project root) to populate section 6:\n\n")
  cat("```bash\n")
  cat(sprintf("bash scripts/build_per_seq_trees.sh %s %s\n", params$batch_dir, params$dataset_date))
  cat("```\n\n")
  cat("Short-region alignment inspection is still available in section 0 under **Junction alignment (focused viewer)**.\n\n")
} else { # tree or alignment files are available
for (i in seq_along(query_ids)) {
  query_id <- query_ids[i]
  query_id_norm <- query_ids_normalized[i]
  cat(paste0("\n## ", query_id, "\n\n"))

  tree_dir <- file.path(trees_dir, query_id_norm)
  # Prefer BLAST-only alignment if available, otherwise fall back to full alignment
  aln_file <- if (file.exists(file.path(tree_dir, "aligned_trimmed_blast_only.fa"))) {
    file.path(tree_dir, "aligned_trimmed_blast_only.fa")
  } else {
    file.path(tree_dir, paste0(query_id_norm, ".aligned_trimmed.fasta"))
  }
  treefile <- file.path(tree_dir, "tree.treefile")
  tr <- NULL

  # Read alignment once — used by both SNP and MSA sections
  seqs_raw <- if (file.exists(aln_file)) {
    tryCatch(ape::read.dna(aln_file, format = "fasta", as.character = TRUE),
             error = function(e) NULL)
  } else NULL
  # The saved .aligned.fasta/.aligned_trimmed.fasta files mark the query header
  # with a trailing " [QUERY]" for readability — strip it back off so all the
  # exact-match comparisons against query_id below keep working unchanged.
  if (!is.null(seqs_raw)) {
    rownames(seqs_raw) <- sub(" \\[QUERY\\]$", "", rownames(seqs_raw))
  }
  # Defensive: duplicate FASTA headers (e.g. from an older alignment generated
  # before the batch FASTA duplicate-header guard) would otherwise crash
  # data.frame() further down with "duplicate row.names". Keep the first
  # occurrence's name intact (so query_id matching still works) and suffix
  # any repeats instead of failing the whole report.
  if (!is.null(seqs_raw) && anyDuplicated(rownames(seqs_raw)) > 0) {
    dup_ids <- unique(rownames(seqs_raw)[duplicated(rownames(seqs_raw))])
    cat(sprintf("*Warning: duplicate sequence header(s) in alignment (%s) — extra copies were renamed for display.*\n\n",
                paste(dup_ids, collapse = ", ")))
    rownames(seqs_raw) <- make.unique(rownames(seqs_raw))
  }

  # ── 1. Phylogenetic tree ────────────────────────────────────────────────────
  cat("### Phylogenetic tree\n\n")
  if (!file.exists(treefile)) {
    cat("*Tree not available — run `build_per_seq_trees.sh` first.*\n\n")
  } else {
    tr <- tryCatch(read.tree(treefile), error = function(e) NULL)
    if (is.null(tr)) {
      cat("*Could not parse tree.*\n\n")
    } else {
      tr <- phangorn::midpoint(tr)

      # If one/few branches are extreme outliers, branch-length plotting can
      # visually collapse the rest of the tree. In that case, switch to a
      # topology-only view for readability.
      # The ratio alone is misleading for tight clusters (near-zero median
      # branch length inflates the ratio even when the longest branch is
      # small in absolute terms), so also require a genuinely large absolute
      # branch length before falling back to the cladogram view.
      edge_len <- tr$edge.length
      edge_len <- edge_len[is.finite(edge_len) & !is.na(edge_len) & edge_len > 0]
      long_branch_ratio <- if (length(edge_len) > 0) {
        max(edge_len) / max(median(edge_len), 1e-6)
      } else {
        1
      }
      max_edge_len <- if (length(edge_len) > 0) max(edge_len) else 0
      # Disabled for now — always render as a phylogram with real branch lengths.
      # use_cladogram_view <- long_branch_ratio > 25 && max_edge_len > 0.05
      use_cladogram_view <- FALSE

tip_info <- tibble(id = tr$tip.label) %>%
  left_join(
    ref_meta %>%
      select(id_norm, genotype, variant),
    by = c("id" = "id_norm")
  ) %>%
  left_join(
    query_lp %>%
      dplyr::rename(
        id = seqName,
        lineage_phylo_q = lineage_phylo
      ),
    by = "id"
  ) %>%
  left_join(
    community_meta,
    by = "id"
  ) %>%
  mutate(
    lineage_phylo = coalesce(
      lineage_phylo_q,
      community_lineage_phylo
    ),

    genotype = coalesce(
      genotype,
      community_clade
    ),

    is_query = id == query_id,

    tip_label = case_when(
      !is.na(genotype) &
        !is.na(variant) &
        !is.na(lineage_phylo) ~
        paste0(
          id, "  [", genotype, "  ", variant,
          "  /  ", lineage_phylo, "]"
        ),

      !is.na(genotype) &
        !is.na(lineage_phylo) ~
        paste0(
          id, "  [", genotype,
          "  /  ", lineage_phylo, "]"
        ),

      !is.na(genotype) &
        !is.na(variant) ~
        paste0(
          id, "  [", genotype,
          "  ", variant, "]"
        ),

      !is.na(genotype) ~
        paste0(id, "  [", genotype, "]"),

      !is.na(variant) &
        !is.na(lineage_phylo) ~
        paste0(
          id, "  [", variant,
          "  /  ", lineage_phylo, "]"
        ),

      !is.na(lineage_phylo) ~
        paste0(id, "  [", lineage_phylo, "]"),

      !is.na(variant) ~
        paste0(id, "  [", variant, "]"),

      !is.na(community_clade) ~
        paste0(id, "  [", community_clade, "]"),

      TRUE ~ id
    ),

    tip_color = case_when(
      is_query ~ "Query",
      !is.na(variant) & variant != "" ~ variant,
      !is.na(genotype) & genotype != "NA" ~ genotype,
      !is.na(community_clade) ~ community_clade,
      TRUE ~ "Unknown"
    )
  )

      all_colors <- unique(tip_info$tip_color)
      ref_colors <- setdiff(all_colors, "Query")
      pal <- setNames(
        c("#d62728", scales::hue_pal()(length(ref_colors))),
        c("Query", ref_colors)
      )

      p <- if (use_cladogram_view) {
        ggtree(
          tr,
          layout = "fan",
          open.angle = 15,
          branch.length = "none"
        ) %<+% tip_info +
          geom_tippoint(aes(color = tip_color), size = 3) +
          geom_tiplab(aes(label = tip_label), size = 2.6, align = TRUE, linetype = NA, offset = 0.15) +
          scale_color_manual(values = pal, name = "Variant") +
          guides(color = guide_legend(ncol = 2)) +
          theme_void() +
          theme(legend.position = "bottom", plot.margin = margin(5, 5, 5, 5))
      } else {
        ggtree(
          tr,
          layout = "rectangular",
          branch.length = "branch.length"
        ) %<+% tip_info +
          geom_tippoint(aes(color = tip_color), size = 3) +
          geom_tiplab(aes(label = tip_label), size = 2.8, hjust = -0.05) +
          scale_color_manual(values = pal, name = "Variant") +
          guides(color = guide_legend(ncol = 2)) +
          theme_tree2() +
          theme(legend.position = "bottom", plot.margin = margin(5, 5, 5, 5)) +
          hexpand(.6)
      }

      bs_df <- p$data %>%
        filter(!isTip, !is.na(label)) %>%
        mutate(bs_val = suppressWarnings(as.numeric(label))) %>%
        filter(!is.na(bs_val), bs_val >= 50)
      if (nrow(bs_df) > 0) {
        p <- p + geom_text(data = bs_df, aes(x = x, y = y, label = round(bs_val)),
                           inherit.aes = FALSE,
                           size = 2.5, hjust = 1.15, vjust = -0.4, color = "grey35")
      }

      # Persist to output/trees/<seqName>/tree.png (not a tempfile) so it survives after report generation
      tree_png_path <- file.path(tree_dir, "tree.png")
      ggsave(tree_png_path, plot = p, width = 18, height = 9, dpi = 150)
      .embed_img(tree_png_path)
      if (use_cladogram_view) {
        cat("*Display note:* Large branch-length outliers detected. Tree is shown as a topology-only cladogram for readability; branch lengths are still preserved in `tree.treefile`.\n\n")
      }
      cat("\n")

meta_tbl <- tip_info %>%
  arrange(desc(is_query), tip_color, id) %>%
  select(
    ID = id,
    Variant = variant,
    Lineage_phylo = lineage_phylo,
    Genotype = genotype
  ) %>%
  mutate(
    across(
      c(Variant, Lineage_phylo, Genotype),
      ~ if_else(
        is.na(.x) | .x %in% c("NA", ""),
        "—",
        as.character(.x)
      )
    )
  )

      cat(kbl(meta_tbl, caption = paste("Sequences in tree:", query_id)) %>%
            kable_styling(bootstrap_options = c("striped", "condensed"),
                          full_width = FALSE, font_size = 11) %>%
            as.character())
      cat("\n\n")
    }
  }

  # ── 2. SNP distance heatmap ─────────────────────────────────────────────────
  cat("### SNP distance heatmap\n\n")
  
  # Use alignment content as-is.
  # When aligned_trimmed_blast_only.fa exists, this is already query + selected BLAST neighbors.
  # For older batches without BLAST-only alignment we keep existing alignment content.
  seqs_heatmap <- if (!is.null(seqs_raw) && nrow(seqs_raw) > 0) {
    if (nrow(seqs_raw) >= 2) seqs_raw else NULL
  } else {
    NULL
  }
  
  if (is.null(seqs_heatmap) || nrow(seqs_heatmap) < 2) {
    cat("*Alignment not available or too few sequences.*\n\n")
  } else {
    n_tips  <- nrow(seqs_heatmap)
    ids     <- rownames(seqs_heatmap)
    snp_mat <- matrix(0L, n_tips, n_tips, dimnames = list(ids, ids))
    for (i in seq_len(n_tips - 1)) {
      for (j in (i + 1):n_tips) {
        s1 <- seqs_heatmap[i, ]; s2 <- seqs_heatmap[j, ]
        # Count mismatches on all non-gap aligned positions.
        # This keeps SNP distance closer to BLAST mismatch behavior.
        valid <- s1 != "-" & s2 != "-"
        d <- sum(s1[valid] != s2[valid])
        snp_mat[i, j] <- snp_mat[j, i] <- d
      }
    }

    ref_meta_hm <- if (
  nrow(meta) > 0 &&
  all(c("id", "variant") %in% names(meta))
) {
  meta %>%
    transmute(
      id = as.character(id),
      hm_variant = as.character(variant),
      hm_genotype = if ("genotype" %in% names(meta)) {
        as.character(genotype)
      } else {
        NA_character_
      }
    )
} else {
  tibble(
    id = character(),
    hm_variant = character(),
    hm_genotype = character()
  )
}


    lbl_tbl <- tibble(id = ids) %>%
      left_join(ref_meta_hm, by = "id") %>%
      left_join(community_meta, by = "id") %>%
      mutate(
        lineage_phylo = community_lineage_phylo,
        variant       = hm_variant,
        genotype      = coalesce(harmonise_genotype(hm_genotype), community_clade),  # Use same as lineage for genotype field
        lineage_phylo = if_else(is.na(lineage_phylo) | lineage_phylo == "",
                                NA_character_, lineage_phylo),
        # Shorten long FASTA headers: extract accession + isolate/strain name
        # Handles both "isolate XXXX" and "strain: XXXX" formats
        # E.g., "KX151409.1 Hepatovirus A isolate h737 ..." → "KX151409.1 (h737)"
        # E.g., "LC035017.1 Hepatitis A virus gene for polyprotein, partial cds, strain: 1404-20055-WakymC-4HA42" → "LC035017.1 (1404-20055-WakymC-4HA42)"
        short_id = {
          accession <- str_extract(id, "^[^\\s]+")
          # Extract isolate/strain name: try "isolate" pattern, then "strain:" pattern
          isolate_m1 <- str_extract(id, "(?<= isolate )[^ ,]+")
          isolate_m2 <- str_extract(id, "(?<=strain: )[^ ,]+")
          isolate_part <- coalesce(isolate_m1, isolate_m2)
          
          if_else(!is.na(isolate_part),
                 paste0(accession, " (", isolate_part, ")"),
                 accession)
        },
        display = case_when(
  id == query_id ~
    paste0(short_id, "  \u2605"),

  !is.na(genotype) &
    !is.na(variant) &
    !is.na(lineage_phylo) ~
    paste0(
      short_id, "  [", genotype, "  ",
      variant, "  /  ", lineage_phylo, "]"
    ),

  !is.na(genotype) &
    !is.na(lineage_phylo) ~
    paste0(
      short_id, "  [", genotype,
      "  /  ", lineage_phylo, "]"
    ),

  !is.na(genotype) &
    !is.na(variant) ~
    paste0(
      short_id, "  [", genotype,
      "  ", variant, "]"
    ),

  !is.na(genotype) ~
    paste0(short_id, "  [", genotype, "]"),

  !is.na(variant) &
    !is.na(lineage_phylo) ~
    paste0(
      short_id, "  [", variant,
      "  /  ", lineage_phylo, "]"
    ),

  !is.na(variant) ~
    paste0(short_id, "  [", variant, "]"),

  !is.na(lineage_phylo) ~
    paste0(short_id, "  [", lineage_phylo, "]"),

  TRUE ~ short_id
)
      )
    lbl_map <- setNames(lbl_tbl$display, lbl_tbl$id)
    rownames(snp_mat) <- lbl_map[rownames(snp_mat)]
    colnames(snp_mat) <- lbl_map[colnames(snp_mat)]

    query_lbl <- lbl_map[query_id]
    ann_row <- data.frame(
      Type = ifelse(rownames(snp_mat) == query_lbl, "Query", "Reference"),
      row.names = rownames(snp_mat)
    )
    ann_colors <- list(Type = c(Query = "#d62728", Reference = "#aec7e8"))

    # Adaptive sizing: base cell size on number of sequences
    # Ensures readable labels and numbers even with many sequences
    cell_size <- if (n_tips <= 10) 60 else if (n_tips <= 20) 45 else 30
    fontsize_main <- if (n_tips <= 10) 15 else if (n_tips <= 20) 14 else 13
    fontsize_num <- if (n_tips <= 10) 12 else if (n_tips <= 20) 11 else 10
    fontsize_row <- if (n_tips <= 15) 13 else 11
    
    # Calculate figure size with minimum/maximum constraints
    px <- min(18, max(10, n_tips * (cell_size / 35)))
    
    # Continuous blue palette for all SNP distances
    color_func <- colorRampPalette(c("#f7fbff", "#deebf7", "#9ecae1", "#3182bd", "#08519c", "#08306b"))(100)
    
    .tmp_heat <- tempfile(fileext = ".png")
    .ph_obj <- pheatmap(
      snp_mat,
      # Color palette: continuous blue
      color            = color_func,
      clustering_distance_rows = as.dist(snp_mat),
      clustering_distance_cols = as.dist(snp_mat),
      clustering_method        = "average",
      display_numbers          = TRUE,
      number_format            = "%d",
      number_color             = "black",      # Improved contrast
      fontsize                 = fontsize_main,
      fontsize_number          = fontsize_num,
      fontsize_row             = fontsize_row,
      annotation_row           = ann_row,
      annotation_colors        = ann_colors,
      legend                   = TRUE,
      main                     = paste("Pairwise SNP distances \u2014", query_id),
      treeheight_row           = 100,
      treeheight_col           = 100,
      border_color             = "#cccccc",    # Subtle borders for better separation
      gaps_row                 = which(rownames(snp_mat) == query_lbl),
      gaps_col                 = which(colnames(snp_mat) == query_lbl),
      silent                   = TRUE
    )
    ggsave(.tmp_heat, plot = .ph_obj$gtable, width = px, height = px, dpi = 150, limitsize = FALSE)
    .embed_img(.tmp_heat)
    file.remove(.tmp_heat)
    
    # Save SNP distance matrix as TSV for external analysis
    snp_tsv <- file.path(out_dir, "snp_matrices", paste0(query_id, "_snp_distances.tsv"))
    dir.create(dirname(snp_tsv), showWarnings = FALSE, recursive = TRUE)
    
    # Convert matrix to data frame with sequence IDs in first column
    snp_df <- as.data.frame(snp_mat)
    snp_df <- tibble::rownames_to_column(snp_df, var = "Sequence_1")
    names(snp_df)[-1] <- rownames(snp_mat)  # Ensure column names are sequence labels
    
    write_tsv(snp_df, snp_tsv)
    cat("[Download SNP distance matrix](snp_matrices/", query_id, "_snp_distances.tsv)\n\n", sep = "")
    
    cat("\n\n")
  } # end SNP heatmap display


---

# 5. Variant naming

```{r variant-naming, results='asis'}
# Extract year from dataset_date parameter (format: YYYY-MM-DD)
dataset_year <- as.numeric(substr(params$dataset_date, 1, 4))
# Year for variant naming (might change at new year)
variant_year <- as.numeric(format(Sys.Date(), "%Y"))

# Check latest variant in database by NOR-YYYY pattern
if (nrow(meta) > 0 && "variant" %in% names(meta)) {
  db_lineages <- meta %>%
    filter(
      !is.na(variant), 
      variant !="",
      grepl("^NOR-", variant)
      ) %>%
    pull(variant) %>%
    unique()
  
  current_year_lineages <- db_lineages[grepl(paste0("^NOR-", variant_year), db_lineages)]
  if (length(current_year_lineages) > 0) {
    # Extract version numbers (V1, V2, etc.) and sort numerically
    # Pattern: NOR-YYYY-VNN[a-z]? → extract NN part
    base_versions <- as.numeric(gsub("^NOR-[0-9]+-V([0-9]+).*", "\\1", current_year_lineages))
    latest_v <- max(base_versions, na.rm = TRUE)
    next_v <- latest_v + 1
  } else {
    latest_v <- 0
    next_v <- 1
  }
  
  # Show recent variants grouped by base version with b/c suffixes
  # Sort by year descending, then version number numerically
  cat("**Recent lineages in database:**\n\n")
  
  # Parse lineages to extract base version and suffix
  variant_parsed <- tibble(variant = unique(db_lineages)) %>%
    filter(grepl("^NOR-", variant)) %>%
    mutate(
      # Extract year from variant (NOR-YYYY-Vx → YYYY)
      year = as.numeric(str_extract(variant, "NOR-([0-9]+)", group = 1)),
      # Extract version full (V1, V1b, etc.)
      version_full = str_extract(variant, "V[0-9]+[a-z]*"),
      # Extract base version number
      base_v = as.numeric(str_extract(version_full, "[0-9]+"))
    ) %>%
    distinct(year, base_v, .keep_all = TRUE) %>%  # Keep only unique year+base_v combinations
    arrange(desc(year), desc(base_v)) %>%
    head(10) %>%  # Get top 10 most recent base versions
    arrange(year, base_v)  # Sort for display: oldest to newest
  
  # Group and display by year, then by base version (oldest to newest for display)
  for (yr in unique(sort(variant_parsed$year))) {
    year_data <- variant_parsed %>% filter(year == yr)
    cat("**NOR-", yr, ":**", sep = "")
    # Collect all variants for each base_v and display on same line
    variants_by_base <- list()
    for (bv in unique(sort(year_data$base_v))) {
      variants <- year_data %>%
        filter(base_v == bv) %>%
        pull(variant) %>%
        sort()
      variants_by_base[[as.character(bv)]] <- variants
    }
    # Display: V1 [V1b] format or V1 [no variants]
    display_str <- paste(sapply(names(variants_by_base), function(v) {
      base <- paste0("NOR-", yr, "-V", v)
      all_vars <- variants_by_base[[v]]
      suffixed <- all_vars[grepl("[a-z]$", all_vars)]
      if (length(suffixed) > 0) {
        paste0(base, " [", paste(sub(base, "", suffixed), collapse = ", "), "]")
      } else {
        base
      }
    }), collapse = ", ")
    cat(" ", display_str, "\n", sep = "")
  }
  cat("\n")
  
  cat(sprintf("**Latest NOR-%d variant: V%d, next candidate: V%d**\n\n", variant_year, latest_v, next_v))
} else {
  cat("*Unable to determine variant naming scheme — no variant data in reference database.*\n\n")
}

# Build variant suggestions table with smart b/c suffix suggestion
if (nrow(summary_tbl) > 0 && nrow(meta) > 0 && "variant" %in% names(meta)) {
  # Function to suggest next available suffix (b, c, d, etc.)
  suggest_suffix <- function(base_lineage, existing_lineages) {
    # Get all variants of this variant in the database
    variants <- existing_lineages[grepl(paste0("^", base_lineage, "[a-z]?$"), existing_lineages)]
    
    if (length(variants) == 0) {
      # No variant exists, suggest 'b'
      return("b")
    }
    
    # Extract which suffixes exist
    suffixes <- str_extract(variants, "[a-z]$")
    suffixes <- suffixes[!is.na(suffixes)]
    
    n_suffixes <- length(suffixes)
    if (n_suffixes == 0) {
      # Base exists without suffixes, next is 'b'
      return("b")
    }
    
    # Suggest next letter after highest existing suffix
    if (n_suffixes > 0) {
      highest_suffix <- sort(suffixes, decreasing = TRUE)[1]
      highest_code <- as.numeric(charToRaw(highest_suffix))
      next_suffix <- rawToChar(as.raw(highest_code + 1))
      return(next_suffix)
    }
    
    return("b")
  }
  

  all_db_lineages <- meta %>%
  filter(!is.na(variant), variant != "") %>%
  pull(variant) %>%
  unique()
  
  variant_suggestions <- summary_tbl %>%
    select(seqName, clade, blast_genotype, blast_variant, blast_mismatch, sseqid) %>%
    mutate(
      lineage_suggestion = case_when(
        !is.na(blast_mismatch) & blast_mismatch == 0 & !is.na(blast_variant) ~ blast_variant,
        !is.na(blast_mismatch) & blast_mismatch == 1 & !is.na(blast_variant) ~ {
          suggested_suffix <- suggest_suffix(blast_variant, all_db_lineages)
          paste0(blast_variant, suggested_suffix, " (1 SNP diff)")
        },
        !is.na(blast_mismatch) & blast_mismatch == 2 & !is.na(blast_variant) ~ {
          suggested_suffix <- suggest_suffix(blast_variant, all_db_lineages)
          paste0(blast_variant, suggested_suffix, " (2 SNP diff)")
        },
        is.na(blast_variant) ~ "No BLAST hit — needs manual assignment",
        TRUE ~ "Review: >2 SNP difference from best hit"
      ),
      notes = case_when(
        !is.na(blast_mismatch) & blast_mismatch == 0 ~ "Identical to database sequence",
        !is.na(blast_mismatch) & blast_mismatch <= 2 ~ paste("Close to", sseqid),
        TRUE ~ ""
      )
    ) %>%
    select(
      Sequence = seqName,
      Genotype = clade,
      `BLAST hit` = sseqid,
      `SNPs` = blast_mismatch,
      `Hit variant` = blast_variant,
      `Suggested variant` = lineage_suggestion,
      Notes = notes
    )
  
  cat(variant_suggestions %>%
        kbl(escape = FALSE) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
        as.character())
} else {
  cat("*No sequences for variant naming.*\n\n")
}
```

**Variant naming protocol:**
- **Identical to database (0 SNPs):** Assign same variant as BLAST hit
- **1-2 SNPs from database sequence:** Check database for existing b/c suffixes, suggest next available (e.g., b if none exist, c if b exists)
- **>2 SNPs:** Manual review required; may warrant new variant number

---

# 6. Batch QC

```{r batch-qc, results='asis'}
# Get batch sample names (first token, trimmed)
batch_sample_names <- word(query_ids_normalized, 1)
batch_seqs_in_alignment <- batch_sample_names[batch_sample_names %in% query_ids_normalized]

if (length(batch_seqs_in_alignment) < 2) {
  cat("*Batch QC analysis requires at least 2 samples in the batch. Skipping batch phylogenetic analysis.*\n\n")
} else {
  # Try to auto-generate batch analysis if files don't exist
  batch_tree_file <- file.path(trees_dir, "batch", "tree.treefile")
  batch_aln_file_aligned <- file.path(trees_dir, "batch", "batch.aligned.fasta")
  batch_tree_available <- file.exists(batch_tree_file) && file.info(batch_tree_file)$size > 0
  batch_aln_available <- file.exists(batch_aln_file_aligned) && file.info(batch_aln_file_aligned)$size > 0
  
  # If batch files missing, try to run build_batch_analysis.sh
  if (!batch_tree_available || !batch_aln_available) {
    build_script <- file.path(proj_dir, "scripts", "build_batch_analysis.sh")
    if (file.exists(build_script)) {
      cat("*Generating batch-specific phylogenetic analysis...*\n\n")
      system(sprintf("bash '%s' '%s' '%s' '%s'",
                     build_script, params$batch_dir, params$dataset_date, params$out_base),
             ignore.stdout = FALSE, ignore.stderr = FALSE)
      # Re-check if files were created
      batch_tree_available <- file.exists(batch_tree_file) && file.info(batch_tree_file)$size > 0
      batch_aln_available <- file.exists(batch_aln_file_aligned) && file.info(batch_aln_file_aligned)$size > 0
    } # end auto-generate batch analysis
  } # end batch file availability check
  cat("## Batch phylogenetic tree\n\n")
  
  batch_tree_obj <- NULL
  if (batch_tree_available) {
    batch_tree_obj <- tryCatch(read.tree(batch_tree_file), error = function(e) NULL)
  }
  
  if (is.null(batch_tree_obj)) {
    # Tree not available - provide instructions
    cat("*Batch phylogenetic tree not available.*\n\n")
    cat("**To generate batch-specific phylogenetic tree and SNP matrix, run:**\n\n")
    cat("```bash\n")
    cat("bash scripts/build_batch_analysis.sh ")
    cat(sprintf("'%s' '%s' '%s'\n", params$batch_dir, params$dataset_date, params$out_base))
    cat("```\n\n")
    cat("This script will:\n")
    cat("1. Extract batch sequences from per-sequence analyses\n")
    cat("2. Align sequences with MAFFT\n")
    cat("3. Infer phylogenetic tree with IQ-TREE\n")
    cat("4. Compute SNP distance matrix\n\n")
    cat("After running, re-generate this report to include batch phylogenetics.\n\n")
  }  else {
  # end batch tree availability check 
    # We have the batch tree - visualize it
    batch_tree_obj <- phangorn::midpoint(batch_tree_obj)
    
    # Metadata for batch tips
    batch_tip_info <- tibble(id = batch_tree_obj$tip.label) %>%
      left_join(summary_tbl %>% select(seqName, clade, lineage_phylo, blast_variant) %>% 
                mutate(id = word(seqName, 1)) %>%
                select(id, clade, lineage_phylo, blast_variant),
                by = "id") %>%
      mutate(
        is_query = id %in% batch_seqs_in_alignment,
        tip_label = case_when(
          !is.na(clade) & !is.na(lineage_phylo) & !is.na(blast_variant) ~ paste0(id, "  [", clade, " / ", lineage_phylo, " / ", blast_variant, "]"),
          !is.na(clade) ~ paste0(id, "  [", clade, "]"),
          !is.na(lineage_phylo) ~ paste0(id, "  [", lineage_phylo, "]"),
          TRUE ~ id
        )
      )
    
    # Plot batch tree
    p_batch <- ggtree(batch_tree_obj, layout = "rectangular", branch.length = "branch.length") %<+% batch_tip_info +
      geom_tippoint(aes(color = clade), size = 3) +
      geom_tiplab(aes(label = tip_label), size = 2.8, hjust = -0.05) +
      scale_color_manual(values = scales::hue_pal()(length(unique(batch_tip_info$clade))), 
                        name = "Clade", na.value = "grey50") +
      theme_tree2() +
      theme(legend.position = "bottom") +
      hexpand(.6) +
      labs(title = sprintf("Batch %s — Phylogenetic tree (batch samples only)", batch_name),
           subtitle = "Phylogenetic relationships among batch sequences")
    
    .tmp_batch_tree <- tempfile(fileext = ".png")
    ggsave(.tmp_batch_tree, plot = p_batch, width = 14, height = 8, dpi = 150)
    cat(sprintf('<img src="%s" style="width:100%%;"/>\n\n', knitr::image_uri(.tmp_batch_tree)))
    file.remove(.tmp_batch_tree)
  } # end batch tree visualization
  
  cat("## Batch SNP distance heatmap\n\n")
  
  batch_snp_file <- file.path(params$out_base, "snp_matrices", "batch_snp_distances.tsv")
  batch_snp_mat <- NULL
  
  # Option 1: Use pre-computed SNP matrix from build_batch_analysis.sh
  if (file.exists(batch_snp_file)) {
    batch_snp_mat <- tryCatch({
      mat <- read.delim(batch_snp_file, row.names = 1, check.names = FALSE)
      as.matrix(mat)
    }, error = function(e) NULL)
  }
  
  # Option 2: Build from per-sequence blast-only alignments (fallback)
  if (is.null(batch_snp_mat)) {
    batch_seqs_data <- tibble()
    
    for (seq_id in batch_seqs_in_alignment) {
      seq_aln_file <- file.path(trees_dir, seq_id, "aligned_trimmed_blast_only.fa")
      if (file.exists(seq_aln_file)) {
        tryCatch({
          aln <- ape::read.dna(seq_aln_file, format = "fasta", as.character = TRUE)
          if (nrow(aln) > 0) {
            # Get first row (query sequence)
            query_seq <- aln[1, , drop = FALSE]
            rownames(query_seq) <- seq_id
            batch_seqs_data <- rbind(batch_seqs_data, as_tibble(query_seq, rownames = "id"))
          }
        }, error = function(e) NULL)
      }
    }
    
    # Compute SNP distance matrix if we have sequences
    if (nrow(batch_seqs_data) >= 2) {
      batch_seqs_mat <- as.matrix(batch_seqs_data %>% select(-id))
      rownames(batch_seqs_mat) <- batch_seqs_data$id
      
      n_batch <- nrow(batch_seqs_mat)
      batch_snp_mat <- matrix(0L, n_batch, n_batch, dimnames = list(rownames(batch_seqs_mat), rownames(batch_seqs_mat)))
      
      for (i in seq_len(n_batch - 1)) {
        for (j in (i + 1):n_batch) {
          s1 <- batch_seqs_mat[i, ]
          s2 <- batch_seqs_mat[j, ]
          valid <- s1 != "-" & s2 != "-" & !is.na(s1) & !is.na(s2)
          d <- sum(s1[valid] != s2[valid])
          batch_snp_mat[i, j] <- batch_snp_mat[j, i] <- d
        }
      }
    }
  }
  
  # Display SNP heatmap if we have matrix
  if (!is.null(batch_snp_mat)) {
    # Defensive: duplicate sequence names would otherwise crash data.frame()
    # with "duplicate row.names" further down.
    if (anyDuplicated(rownames(batch_snp_mat)) > 0) {
      rownames(batch_snp_mat) <- make.unique(rownames(batch_snp_mat))
      colnames(batch_snp_mat) <- make.unique(colnames(batch_snp_mat))
    }
    batch_lbl <- rownames(batch_snp_mat)
    batch_ann_row <- data.frame(
      Type = rep("Batch", nrow(batch_snp_mat)),
      row.names = batch_lbl
    )
    batch_ann_colors <- list(Type = c(Batch = "#aec7e8"))
    
    .tmp_batch_heat <- tempfile(fileext = ".png")
    .pb_obj <- pheatmap(
      batch_snp_mat,
      color = colorRampPalette(c("#f7fbff", "#deebf7", "#9ecae1", "#3182bd", "#08519c", "#08306b"))(100),
      clustering_distance_rows = as.dist(batch_snp_mat),
      clustering_distance_cols = as.dist(batch_snp_mat),
      clustering_method = "average",
      display_numbers = TRUE,
      number_format = "%d",
      number_color = "black",
      fontsize = 14,
      fontsize_number = 12,
      annotation_row = batch_ann_row,
      annotation_colors = batch_ann_colors,
      main = sprintf("Batch %s — SNP distances", batch_name),
      silent = TRUE
    )
    ggsave(.tmp_batch_heat, plot = .pb_obj$gtable, width = 10, height = 10, dpi = 150)
    cat(sprintf('<img src="%s" style="width:100%%;"/>\n\n', knitr::image_uri(.tmp_batch_heat)))
    file.remove(.tmp_batch_heat)
  } else {
  # end batch SNP heatmap display
    cat("*Batch SNP distance matrix could not be computed. Run the batch analysis script to generate:*\n\n")
    cat("```bash\n")
    cat("bash scripts/build_batch_analysis.sh ")
    cat(sprintf("'%s' '%s' '%s'\n", params$batch_dir, params$dataset_date, params$out_base))
    cat("```\n\n")
  } # end batch SNP fallback
} # end batch size check
```

---

# 7. Outbreak detection

```{r outbreak-detection, results='asis'}
# ── Helper function for displaying images with lightbox ───────────────────────
.embed_img <- function(path) {
  if (!file.exists(path)) {
    cat(sprintf("*(Image file not found: %s)*\n", basename(path)))
    return()
  }
  cat(sprintf(
    '<img src="%s" onclick="document.getElementById(\'tree-lightbox-img\').src=this.src; var lb=document.getElementById(\'tree-lightbox\'); lb.style.display=\'flex\';" style="width:100%%; cursor:zoom-in; border:1px solid #ddd; border-radius:4px; margin-bottom:12px;" title="Click to enlarge"/>\n',
    knitr::image_uri(path)
  ))
}

# Load BLAST results to identify which variants batch sequences match
# Variants come from database variant assignments
blast_data <- read_tsv(
  file.path(params$out_base, "blast_results.tsv"),
  col_names = c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen",
                "qstart", "qend", "sstart", "send", "evalue", "bitscore"),
  col_types = "ccddddiiiidd",
  show_col_types = FALSE
)

# Define robust ID normalization
norm_id <- function(x) {
  x <- iconv(x, to = "UTF-8", sub = "")
  x <- tolower(x)
  gsub("[-_]", "", x, perl = TRUE)
}

# Extract base variant (remove trailing a/b/c/d suffixes)
extract_base_variant <- function(variant) {
  ifelse(is.na(variant), NA_character_, sub("[a-z]$", "", variant))
}

# Identify batch variants:
# For each batch sequence, get best BLAST hit (first row, sorted by score)
# Then look up the variant/lineage from database metadata
# Group by BASE variant (removing sub-variant suffixes like a/b/c)
batch_variants <- blast_data %>%
  arrange(
    qseqid,
    mismatch,
    desc(bitscore)
  ) %>%
  group_by(qseqid) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  select(
    qseqid,
    sseqid
  ) %>%
  mutate(
    sseqid_norm = norm_id(sseqid)
  ) %>%
  left_join(
    meta %>%
      filter(
        !is.na(variant),
        variant != ""
      ) %>%
      mutate(
        id_norm = norm_id(id)
      ) %>%
      select(
        id_norm,
        variant
      ),
    by = c("sseqid_norm" = "id_norm")
  ) %>%
  select(
    seqName = qseqid,
    variant
  ) %>%
  filter(
    !is.na(variant)
  ) %>%
  mutate(
    base_variant = extract_base_variant(variant)
  )
  # Normalize IDs for matching with metadata
  batch_variants <- blast_data %>%
  arrange(
    qseqid,
    mismatch,
    desc(bitscore)
  ) %>%
  group_by(qseqid) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  select(
    qseqid,
    sseqid
  ) %>%
  mutate(
    sseqid_norm = norm_id(sseqid)
  ) %>%
  left_join(
    meta %>%
      filter(
        !is.na(variant),
        variant != ""
      ) %>%
      mutate(
        id_norm = norm_id(id)
      ) %>%
      select(
        id_norm,
        variant
      ),
    by = c("sseqid_norm" = "id_norm")
  ) %>%
  select(
    seqName = qseqid,
    variant
  ) %>%
  filter(
    !is.na(variant)
  ) %>%
  mutate(
    base_variant = extract_base_variant(variant)
  )

# Group by base variant
batch_variant_summary <- batch_variants %>%
  group_by(base_variant) %>%
  summarise(
    n_batch = n(),
    batch_seqs = paste(
      sort(unique(seqName)),
      collapse = ", "
    ),
    .groups = "drop"
  )

# Add database counts for context
if (nrow(meta) > 0 && "variant" %in% names(meta)) {
  db_variant_summary <- meta %>%
    filter(
      !is.na(variant),
      variant != ""
    ) %>%
    mutate(
      base_variant = extract_base_variant(variant)
    ) %>%
    group_by(base_variant) %>%
    summarise(
      n_db = n(),
      .groups = "drop"
    )

  outbreak_candidates <- batch_variant_summary %>%
    left_join(
      db_variant_summary,
      by = "base_variant"
    ) %>%
    mutate(
      n_db = replace_na(n_db, 0L),
      n_total = n_batch + n_db
    ) %>%
    select(
      base_variant,
      n_batch,
      n_db,
      n_total,
      batch_seqs
    )
} else {
  outbreak_candidates <- batch_variant_summary %>%
    mutate(
      n_db = 0L,
      n_total = n_batch
    ) %>%
    select(
      base_variant,
      n_batch,
      n_db,
      n_total,
      batch_seqs
    )
}

if (nrow(outbreak_candidates) == 0) {
  cat("*No variants detected in batch.*\n\n")
} else {
  # Table shows variants with >=2 total sequences; trees require >=3 (built in build_outbreak_trees.R)
  outbreak_candidates_table <- outbreak_candidates %>%
    filter(n_total >= 2)
  outbreak_candidates_display <- outbreak_candidates %>%
    filter(n_total >= 3)
  
  if (nrow(outbreak_candidates_table) == 0) {
    cat("*No lineages with ≥2 total sequences detected in batch.*\n\n")
  } else {
    cat(sprintf("**Variants in batch:** %d variant group(s) with ≥2 sequences\n\n", nrow(outbreak_candidates_table)))
    
    # Summary table 
    cat("**Variant overview (≥2 sequences):**\n\n")
    outbreak_display <- outbreak_candidates_table %>%
      select(
        `Variant` = base_variant,
        `Batch (n)` = n_batch,
        `Database (n)` = n_db,
        `Total` = n_total
      )
    
    cat(outbreak_display %>%
          kbl() %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
        as.character())
    cat("\n\n")
  }
  
  # Create tabset structure for outbreak trees (display pre-generated PNG images)
  # NOTE: {.tabset} must be attached to an actual Markdown header, not bold text,
  # and tab headers must be exactly one level deeper for the tabset to activate.
  cat("## Phylogenetic trees by variant {.tabset}\n\n")
  
  # Check if outbreak_trees directory exists with PNG files
  outbreak_trees_dir <- file.path(params$out_base, "outbreak_trees")
  
  if (dir.exists(outbreak_trees_dir)) {
    png_files <- list.files(outbreak_trees_dir, pattern = "\\.png$", full.names = TRUE)
    png_files <- sort(png_files)  # Sort alphabetically
    
    if (length(png_files) > 0) {
      # Create tabs for each PNG file (only variants with >=3 sequences)
      for (png_path in png_files) {
        variant_name <- sub("\\.png$", "", basename(png_path))
        
        # Get batch and DB sequence counts for this variant (from filtered candidates)
        variant_row <- outbreak_candidates_display %>%
          filter(base_variant == variant_name)
        
        # Skip if not in display list (should have >=3 sequences)
        if (nrow(variant_row) == 0) next
        
        n_batch_seqs <- variant_row$n_batch[1]
        n_db_seqs <- variant_row$n_db[1]
        
        # Create tab header
        cat(sprintf("\n### %s\n\n", variant_name))
        cat(sprintf("**Batch samples:** %d | **Database samples:** %d | **Total:** %d\n\n",
                    n_batch_seqs, n_db_seqs, n_batch_seqs + n_db_seqs))
        
        # Embed pre-generated PNG image
        .embed_img(png_path)
        cat("\n\n")
        
        # Show database sequences with this variant
db_seqs_for_variant <- meta %>%
  filter(!is.na(variant), variant != "") %>%
  mutate(
    base_var = extract_base_variant(variant)
  ) %>%
  filter(base_var == variant_name) %>%
  select(
    id,
    variant,
    genotype,
    any_of("date")
  )

if (nrow(db_seqs_for_variant) > 0) {
  cat("**Database sequences with this variant:**\n\n")

  db_display <- db_seqs_for_variant %>%
    rename(
      ID = id,
      Variant = variant,
      Genotype = genotype
    )

  if ("date" %in% names(db_display)) {
    db_display <- db_display %>%
      rename(Date = date)
  }

  cat(
    db_display %>%
      kbl() %>%
      kable_styling(
        bootstrap_options = c(
          "striped",
          "hover",
          "condensed"
        ),
        full_width = TRUE
      ) %>%
      as.character()
  )

  cat("\n\n")
} # end database sequences display

      } # end for each png_file (outbreak trees)
    } else {
      cat("*Outbreak tree images were not generated. Run:*\n\n")
      cat("```bash\n")
      cat(sprintf("Rscript scripts/build_outbreak_trees.R '%s' '%s' '%s'\n",
                  params$out_base, params$dataset_date, params$dataset_dir))
      cat("```\n\n")
    }
  } else {
    cat("*Outbreak tree images not found. Run the outbreak trees generation step:*\n\n")
    cat("```bash\n")
    cat(sprintf("Rscript scripts/build_outbreak_trees.R '%s' '%s' '%s'\n",
                params$out_base, params$dataset_date, params$dataset_dir))
    cat("```\n\n")
  } # end outbreak_trees_dir check
} # end outbreak detection
```

---

# 8. Conclusions and recommendations

```{r conclusions}
n_seqs        <- nrow(summary_tbl)
n_lp          <- sum(is_lineage(summary_tbl$lineage_phylo), na.rm = TRUE)
n_blast_agree <- sum(summary_tbl$clade_agree_blast == TRUE, na.rm = TRUE)
```

- **`r n_seqs`** sequences analyzed.
- **Lineage resolution:** `r n_lp`/`r n_seqs` sequences assigned a `lineage_phylo` by the lineages dataset.
- **BLAST vs clade agreement:** `r n_blast_agree`/`r sum(!is.na(summary_tbl$clade_agree_blast))` sequences where genotype is available.

---

# 9. Output files

```{r output-files}
tibble(
  File = c(
    file.path(params$out_base, "blast_results.tsv"),
    file.path(params$out_base, "lineages", "nextclade.tsv"),
    file.path(params$out_base, "trees", "<seqName>", "tree.treefile"),
    file.path(params$out_base, "trees", "<seqName>", "tree.png"),
    file.path(params$out_base, "trees", "<seqName>", "<seqName>.aligned_trimmed.fasta"),
    file.path(params$out_base, "trees", "<seqName>", "aligned_trimmed_blast_only.fa"),
    file.path(params$out_base, "snp_matrices", "<seqName>_snp_distances.tsv"),
    file.path(params$out_base, "outbreak_trees", "<Variant>.fasta"),
    file.path(params$out_base, "outbreak_trees", "<Variant>.png"),
    file.path(params$out_base, "summary.tsv"),
    file.path(params$out_base, "comparison.tsv")
  ),
  Description = c(
    "BLAST hits vs local HAV database",
    "NextClade: lineages dataset (clade + lineage_phylo)",
    "IQ-TREE ML tree per query sequence",
    "Rendered phylogenetic tree image (PNG) per query sequence",
    "MAFFT alignment (query + BLAST hits + NextClade community sequences)",
    "MAFFT alignment with BLAST hits only (no NextClade) — used for heatmap in report",
    "Pairwise SNP distance matrix per query sequence (TSV format for external analysis)",
    "Outbreak phylogenetic alignment: batch sequences + database sequences for each variant (FASTA)",
    "Outbreak phylogenetic tree visualization (PNG) per HAV variant",
    "Per-sequence summary table (this report, TSV)",
    "Clade comparison table (this report, TSV)"
  )
) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE)
```

```{r write-outputs}
# Write summary TSV
write_tsv(summary_tbl, file.path(params$out_base, "summary.tsv"))

# Write comparison TSV
clade_comp_out <- clade_comp %>%
  left_join(summary_tbl %>% select(seqName, blast_mismatch = blast_mismatch,
                                    blast_pident, blast_qcovs), by = "seqName")
write_tsv(clade_comp_out, file.path(out_dir, "comparison.tsv"))
```

*Summary written to `r file.path(params$out_base, "summary.tsv")`*  
*Comparison written to `r file.path(params$out_base, "comparison.tsv")`*

---

*Report generated: `r Sys.time()`*  
*Batch: `r batch_name`*  
*Dataset: `r params$dataset_date`*  
